---
title: "Vitamin D levels before and during the COVID-19 pandemic: Data Analysis"
output: html_document
date: "2025-02-01"
author: "Lea Skapetze"
---


```{r setup, include=FALSE}
# Set chunk options
knitr::opts_chunk$set(echo = TRUE)

# Load required libraries
library(lubridate)
library(ggplot2)
library(rlang)
library(tidyr)
library(stringr)
library(stats)
library(purrr)
library(ggsignif)
library(MatchIt)
library(lmtest)
library(sandwich)
library(parallel)
library(broom)
library(cobalt)
library(optmatch)
library(lm.beta)
library(grf)
library(tidyverse)
library(mediation)
library(dplyr)
library(sensemakr)
library(ggrepel)
library(cobalt)
library(readr)
library(xtable)
library(nlme)

# Turn off scientific notations
options(scipen = 999)

# Check versions of packages
R.version.string
packageVersion("grf")
packageVersion("MatchIt")
packageVersion("sensemakr")

# Pretty printer: "p = 5.6 × 10⁻¹⁸" with superscripts
format_p_sci <- function(p, digits = 2) {
  if (is.na(p)) return("p = NA")
  if (p == 0)   return("p < 1 × 10⁻³⁰⁸")  # defensive underflow guard

  s <- formatC(p, format = "e", digits = digits)  # e.g., "5.60e-18"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]
  exp  <- parts[2]

  supers <- c(
    "0"="⁰","1"="¹","2"="²","3"="³","4"="⁴",
    "5"="⁵","6"="⁶","7"="⁷","8"="⁸","9"="⁹",
    "-"="⁻","+"="⁺"
  )
  exp_sup <- paste0(supers[strsplit(exp, "")[[1]]], collapse = "")  # <-- collapse!

  paste0("p = ", base, " × 10", exp_sup)
}


```

## 1. Prepare the data
```{r}
df_unclean <- read.csv("/data/analytics/data_package_archive/HI-01_Covid_VitD/v2/for_costumer/20250604_HI-01_Covid_Study_Data_Package.csv")

unique_patients_count <- df_unclean %>%
  distinct(patient) %>%
  nrow()

cat("Unique patients before cleaning steps:", unique_patients_count, "\n")
```


### 1.1 All measurements
```{r}

df_all_measurements <- read.csv("data.csv")

# Count and calculate percentage of unique patients with gender "U"
gender_u_summary <- df_all_measurements %>%
  dplyr::select(patient, gender) %>%
  distinct() %>%
  summarise(
    count_gender_u = sum(gender == "U"),
    percentage_gender_u = (sum(gender == "U") / n()) * 100
  )

# Print the summary
print(gender_u_summary)

 # GENDER
df_all_measurements <- df_all_measurements %>%
  filter(gender != "U")

# AGE GROUPS

df_all_measurements <- df_all_measurements %>%
  mutate(age_group = case_when(
    age %in% c("[18, 30[", "[30, 40[") ~ "[18, 39]",
    age %in% c("[40, 50[", "[50, 60[") ~ "[40, 59]",
    age == ">=60" ~ "60+",
    TRUE ~ "Unknown" # In case there are entries that don't fit the above categories
  ))

# DATES, PERIODS, SEASONS

# Convert 'cohort_year' and 'test_month' to date format
df_all_measurements$test_date <- with(df_all_measurements, paste(cohort_year, test_month, "01", sep="-"))
df_all_measurements$test_date <- as.Date(df_all_measurements$test_date, format="%Y-%m-%d")

# Create a new column `year_month` combining `cohort_year` and `test_month`
df_all_measurements <- df_all_measurements %>%
  mutate(
    year_month = as.Date(paste(cohort_year, test_month, "01", sep = "-"), format = "%Y-%m-%d")
  )

# Define periods “B” for before the pandemic and “D” for during the pandemic
period1 <- c(as.Date("2018-03-01"), as.Date("2020-02-29"))
period2 <- c(as.Date("2020-03-01"), as.Date("2022-02-28"))

# Assign each entry to a period
assign_period <- function(date) {
  if (date >= period1[1] && date <= period1[2]) {
    return("B")
  } else if (date >= period2[1] && date <= period2[2]) {
    return("D")
  } else {
    return(NA) 
  }
}

# Create a new column for the period
df_all_measurements$period <- as.character(sapply(df_all_measurements$test_date, assign_period))

# Filter out entries that do not belong to either period
df_all_measurements <- df_all_measurements[!is.na(df_all_measurements$period),]

# Define a function to convert test month to season and adjust cohort_year
# The get_season() function adjusts the cohort_year for December by incrementing the year (year + 1), ensuring that December 2019 belongs to Winter 2020
get_season <- function(month, year) {
  if (month %in% c(12)) {
    return(list(season = "Winter", cohort_year = year + 1)) # December moves to the next year's Winter
  } else if (month %in% c(1, 2)) {
    return(list(season = "Winter", cohort_year = year)) # January and February are Winter of the current year 
  } else if (month %in% c(3, 4, 5)) {
    return(list(season = "Spring", cohort_year = year))
  } else if (month %in% c(6, 7, 8)) {
    return(list(season = "Summer", cohort_year = year))
  } else if (month %in% c(9, 10, 11)) {
    return(list(season = "Fall", cohort_year = year))
  } else {
    return(list(season = NA, cohort_year = year)) # Handle unexpected values
  }
}
# For December measurements: In the end the column cohort_year is the reassigned year, and the column year is the original year

# Apply the function to adjust both season and cohort_year
adjusted_season <- mapply(get_season, df_all_measurements$test_month, df_all_measurements$cohort_year, SIMPLIFY = FALSE)
df_all_measurements$season <- factor(sapply(adjusted_season, function(x) x$season), levels = c("Winter", "Spring", "Summer", "Fall"))
df_all_measurements$cohort_year <- sapply(adjusted_season, function(x) x$cohort_year)

# Create ordered season-year by sorting year and season
df_all_measurements <- df_all_measurements %>%
  mutate(
    season_year = paste(cohort_year, season)
  ) %>%
  arrange(cohort_year, season) %>%
  mutate(
    season_year = factor(season_year, levels = unique(season_year), ordered = TRUE)
  )

# VIT D DEFICIENCY STATUS

# Convert the column for value to numeric instead of character 
df_all_measurements <- df_all_measurements %>%
  mutate(value_result = as.numeric(gsub(",", ".", value_result)))

# Filter df_all_measurements for only valid entries in value_result column
df_all_measurements <- df_all_measurements %>% 
  filter(!is.na(value_result))

# Determining deficiency status 
thresholds <- list(
  VD3 = 20  # to be changed for a different param
)

# Function to check deficiency for Vitamin D
get_deficiency_status <- function(value) {
  threshold <- thresholds$VD3 # to be changed for a different param
  if (!is.na(value) && value < threshold) {
    return(1)  # Deficient
  } else {
    return(0)  # Not deficient
  }
}


# Calculation of deficiency status based on numeric value_result column
df_all_measurements <- df_all_measurements %>%
  rowwise() %>%
  mutate(deficient = get_deficiency_status(value_result)) %>%
  ungroup()

# Make sure there are not NAs
df_all_measurements <- df_all_measurements %>% filter(!is.na(value_result), !is.na(period), !is.na(gender), !is.na(age_group))

# Factorize relevant columns for Propensity-Score Matching and Machine Learning
df_all_measurements$gender <- factor(df_all_measurements$gender)
df_all_measurements$age_group <- factor(df_all_measurements$age_group)
df_all_measurements$period <- factor(df_all_measurements$period)
df_all_measurements$test_month <- factor(df_all_measurements$test_month)
df_all_measurements$season <- factor(df_all_measurements$season)


head(df_all_measurements)

# STATISTICS

# Define Z Scores
z_99 <- 2.576
z_95 <- 1.960 

```

### 1.2 First measurement only
```{r}
df <- read.csv("/data/analytics/data_package_archive/HI-01_Covid_VitD/v2/for_costumer/20250604_HI-01_Covid_Study_Data_Package.csv")

# Limit cleaned data frame to only first measurement per patient
df <- df %>%
  arrange(patient, cohort_year, test_month) %>%  
  group_by(patient) %>%                          
  slice_head(n = 1) %>%                         
  ungroup()

# GENDER
df <- df %>%
  filter(gender != "U")

# AGE GROUPS

df <- df %>%
  mutate(age_group = case_when(
    age %in% c("[18, 30[", "[30, 40[") ~ "[18, 39]",
    age %in% c("[40, 50[", "[50, 60[") ~ "[40, 59]",
    age == ">=60" ~ "60+",
    TRUE ~ "Unknown" # In case there are entries that don't fit the above categories
  ))

# DATES, PERIODS, SEASONS

# Convert 'cohort_year' and 'test_month' to date format
df$test_date <- with(df, paste(cohort_year, test_month, "01", sep="-"))
df$test_date <- as.Date(df$test_date, format="%Y-%m-%d")

# Create a new column `year_month` combining `cohort_year` and `test_month`
df <- df %>%
  mutate(
    year_month = as.Date(paste(cohort_year, test_month, "01", sep = "-"), format = "%Y-%m-%d")
  )

# Define periods “B” for before the pandemic and “D” for during the pandemic
period1 <- c(as.Date("2018-03-01"), as.Date("2020-02-29"))
period2 <- c(as.Date("2020-03-01"), as.Date("2022-02-28"))

# Assign each entry to a period
assign_period <- function(date) {
  if (date >= period1[1] && date <= period1[2]) {
    return("B")
  } else if (date >= period2[1] && date <= period2[2]) {
    return("D")
  } else {
    return(NA) 
  }
}

# Create a new column for the period
df$period <- as.character(sapply(df$test_date, assign_period))

# Filter out entries that do not belong to either period
df <- df[!is.na(df$period),]

# Define a function to convert test month to season and adjust cohort_year
# The get_season() function adjusts the cohort_year for December by incrementing the year (year + 1), ensuring that December 2019 belongs to Winter 2020
get_season <- function(month, year) {
  if (month %in% c(12)) {
    return(list(season = "Winter", cohort_year = year + 1)) # December moves to the next year's Winter
  } else if (month %in% c(1, 2)) {
    return(list(season = "Winter", cohort_year = year)) # January and February are Winter of the current year 
  } else if (month %in% c(3, 4, 5)) {
    return(list(season = "Spring", cohort_year = year))
  } else if (month %in% c(6, 7, 8)) {
    return(list(season = "Summer", cohort_year = year))
  } else if (month %in% c(9, 10, 11)) {
    return(list(season = "Fall", cohort_year = year))
  } else {
    return(list(season = NA, cohort_year = year)) # Handle unexpected values
  }
}
# For December measurements: In the end the column cohort_year is the reassigned year, and the column year is the original year

# Apply the function to adjust both season and cohort_year
adjusted_season <- mapply(get_season, df$test_month, df$cohort_year, SIMPLIFY = FALSE)
df$season <- factor(sapply(adjusted_season, function(x) x$season), levels = c("Winter", "Spring", "Summer", "Fall"))
df$cohort_year <- sapply(adjusted_season, function(x) x$cohort_year)

# Create ordered season-year by sorting year and season
df <- df %>%
  mutate(
    season_year = paste(cohort_year, season)
  ) %>%
  arrange(cohort_year, season) %>%
  mutate(
    season_year = factor(season_year, levels = unique(season_year), ordered = TRUE)
  )

# VIT D DEFICIENCY STATUS

# Convert the column for value to numeric instead of character 
df <- df %>%
  mutate(value_result = as.numeric(gsub(",", ".", value_result)))

# Filter df for only valid entries in value_result column
df <- df %>% 
  filter(!is.na(value_result))

# Determining deficiency status 
thresholds <- list(
  VD3 = 20  # to be changed for a different param
)

# Function to check deficiency for Vitamin D
get_deficiency_status <- function(value) {
  threshold <- thresholds$VD3 # to be changed for a different param
  if (!is.na(value) && value < threshold) {
    return(1)  # Deficient
  } else {
    return(0)  # Not deficient
  }
}


# Calculation of deficiency status based on numeric value_result column
df <- df %>%
  rowwise() %>%
  mutate(deficient = get_deficiency_status(value_result)) %>%
  ungroup()

# Make sure there are not NAs
df <- df %>% filter(!is.na(value_result), !is.na(period), !is.na(gender), !is.na(age_group))

# Factorize relevant columns for Propensity-Score Matching and Machine Learning
df$gender <- factor(df$gender)
df$age_group <- factor(df$age_group)
df$period <- factor(df$period)
df$test_month <- factor(df$test_month)
df$season <- factor(df$season)


head(df)

# STATISTICS

# Define Z Scores
z_99 <- 2.576
z_95 <- 1.960 
```


## 2. Descriptive Statistics

```{r}
# Calculate percentage of vitamin D levels > 70 µg/l
outlier_count <- df %>% filter(value_result > 70) %>% nrow()
total_count <- nrow(df)
outlier_percentage <- (outlier_count / total_count) * 100

cat("Number of values > 70 µg/l:", outlier_count, "\n")
cat("Percentage of values > 70 µg/l:", round(outlier_percentage, 2), "%\n")
```


### 2.1 Basic comparison
#### 2.1.1 Sample sizes, counts, mean levels and deficiency rates (unstratified)
```{r}
# Count sample size in period "B" and period "D"
sample_size_period <- df %>% 
  count(period)

sample_size_contingency <- with(df, table(period))
sample_size_p_value <- chisq.test(sample_size_contingency)$p.value

# Count how many entries for each age group in period "B" and "D" (absolute nr and percentage)
age_group_counts <- df %>%
  group_by(period, age_group) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

age_group_contingency <- with(df, table(period, age_group))
age_group_p_value <- chisq.test(age_group_contingency)$p.value

# Count how many entries for each gender group in period "B" and "D" (absolute nr and percentage)
gender_group_counts <- df %>%
  group_by(period, gender) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()


gender_group_contingency <- with(df, table(period, gender))
gender_group_p_value <- chisq.test(gender_group_contingency)$p.value

# Count how many entries for each season group in period "B" and "D" (absolute nr and percentage)
season_counts <- df %>%
  group_by(period, season) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ungroup()

season_contingency <- with(df, table(period, season))
season_counts_p_value <- chisq.test(season_contingency)$p.value

# The mean vitamin D level in period B and D (with standard deviation)
mean_vitamin_d_levels <- df %>%
  group_by(period) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

# Vitamin D deficiency rate for period B and D (in percent)
vitamin_d_deficiency_rate <- df %>%
  group_by(period) %>%
  summarise(
    deficiency_count = sum(deficient),             
    total_count = n(), 
    deficiency_rate = sum(deficient) / n() * 100, 
    .groups = 'drop')

# Printing results
print(sample_size_period)
print(sample_size_p_value)
print(age_group_counts)
print(age_group_p_value)
print(gender_group_counts)
print(gender_group_p_value)
print(season_counts)
print(season_counts_p_value)
print(mean_vitamin_d_levels)
print(vitamin_d_deficiency_rate)

# P Values
## ---- all_pvalues_for_table 1 -----------------------------------------------

# Pretty printer: "p = 5.6 × 10⁻¹⁸" with superscripts
format_p_sci <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("p = NA")
  if (p == 0) return("p < 1 × 10⁻³⁰⁸")
  s <- formatC(p, format = "e", digits = digits)  # e.g., "5.60e-18"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- parts[2]
  supers <- c("0"="⁰","1"="¹","2"="²","3"="³","4"="⁴",
              "5"="⁵","6"="⁶","7"="⁷","8"="⁸","9"="⁹",
              "-"="⁻","+"="⁺")
  exp_sup <- paste0(supers[strsplit(exp, "")[[1]]], collapse = "")
  paste0("p = ", base, " × 10", exp_sup)
}

# Helper: per-level χ² vs period (B/D) for a categorical variable
subgroup_chisq_by_period <- function(data, var_name, period_var = "period") {
  stopifnot(period_var %in% names(data), var_name %in% names(data))
  d <- dplyr::filter(data, !is.na(.data[[var_name]]), !is.na(.data[[period_var]]))
  d <- d %>% dplyr::filter(.data[[period_var]] %in% c("B","D"))
  levs <- if (is.factor(d[[var_name]])) levels(d[[var_name]]) else sort(unique(d[[var_name]]))
  total_B <- sum(d[[period_var]] == "B"); total_D <- sum(d[[period_var]] == "D")

  purrr::map_dfr(levs, function(lvl) {
    tab <- table(d[[period_var]], d[[var_name]] == lvl)
    # ensure rows B/D exist
    if (!all(c("B","D") %in% rownames(tab))) {
      miss <- setdiff(c("B","D"), rownames(tab))
      addm <- matrix(0, nrow = length(miss), ncol = ncol(tab),
                     dimnames = list(miss, colnames(tab)))
      tab <- rbind(tab, addm)[c("B","D"), , drop = FALSE]
    }
    p <- suppressWarnings(chisq.test(tab, correct = FALSE)$p.value)
    n_B <- tab["B","TRUE"]; n_D <- tab["D","TRUE"]
    tibble::tibble(
      variable = var_name,
      level = as.character(lvl),
      n_B = as.integer(n_B),
      pct_B = 100 * n_B / total_B,
      n_D = as.integer(n_D),
      pct_D = 100 * n_D / total_D,
      p_value = p,
      p_formatted = format_p_sci(p)
    )
  }) %>% dplyr::arrange(variable, level)
}

# ---------- GLOBAL p-values ----------
# Global χ² for categorical vars
p_age_global    <- suppressWarnings(chisq.test(table(df$period, df$age_group))$p.value)
p_gender_global <- suppressWarnings(chisq.test(table(df$period, df$gender))$p.value)
p_season_global <- suppressWarnings(chisq.test(table(df$period, df$season))$p.value)

# Welch's t-test for mean vitamin D level (B vs D)
p_mean_global <- t.test(value_result ~ period, data = df, var.equal = FALSE)$p.value

# χ² for deficiency rate (deficient vs not, B vs D)
p_def_global  <- suppressWarnings(chisq.test(table(df$period, df$deficient))$p.value)

# ---------- PER-SUBGROUP p-values ----------
p_age_levels    <- subgroup_chisq_by_period(df, "age_group")
p_gender_levels <- subgroup_chisq_by_period(df, "gender")
p_season_levels <- subgroup_chisq_by_period(df, "season")

# ---------- Collect everything table-ready ----------
pvals_global_tbl <- tibble::tibble(
  row_label = c(
    "Age group (global)",
    "Gender (global)",
    "Season (global)",
    "Vitamin D levels (Welch t-test)",
    "Vitamin D deficiency rate"
  ),
  p_value = c(p_age_global, p_gender_global, p_season_global, p_mean_global, p_def_global),
  p_formatted = purrr::map_chr(p_value, format_p_sci)
)

# For per-row entries, we keep the actual level strings from your data
pvals_subgroups_tbl <- dplyr::bind_rows(
  p_age_levels    %>% dplyr::mutate(row_label = paste0("Age: ", level)) %>% dplyr::select(row_label, p_value, p_formatted),
  p_gender_levels %>% dplyr::mutate(row_label = paste0("Gender: ", level)) %>% dplyr::select(row_label, p_value, p_formatted),
  p_season_levels %>% dplyr::mutate(row_label = paste0("Season: ", level)) %>% dplyr::select(row_label, p_value, p_formatted)
)

# Final combined table of p-values (global + per-subgroup)
pvalues_for_latex <- dplyr::bind_rows(
  pvals_global_tbl,
  pvals_subgroups_tbl
)

# Inspect
pvals_global_tbl
pvals_subgroups_tbl
pvalues_for_latex

p_age_global_fmt    <- format_p_sci(p_age_global)
p_gender_global_fmt <- format_p_sci(p_gender_global)
p_season_global_fmt <- format_p_sci(p_season_global)
p_mean_global_fmt   <- format_p_sci(p_mean_global)
p_def_global_fmt    <- format_p_sci(p_def_global)

cat(
  "\nGLOBAL p-values:\n",
  "Age group: ", p_age_global_fmt, "\n",
  "Gender: ", p_gender_global_fmt, "\n",
  "Season: ", p_season_global_fmt, "\n",
  "Mean vitamin D: ", p_mean_global_fmt, "\n",
  "Deficiency rate: ", p_def_global_fmt, "\n",
  sep = ""
)


```

#### 2.1.2 Effect size (Cohen's d) - Mean vitamin D levels
```{r}
# Means and standard deviations
mean_b <- mean(df$value_result[df$period == "B"])
mean_d <- mean(df$value_result[df$period == "D"])
sd_b <- sd(df$value_result[df$period == "B"])
sd_d <- sd(df$value_result[df$period == "D"])

# Pooled SD
n1 <- sum(df$period == "B")
n2 <- sum(df$period == "D")
sd_pooled <- sqrt(((n1 - 1) * sd_b^2 + (n2 - 1) * sd_d^2) / (n1 + n2 - 2))

# Cohen's d (flipped sign for "During - Before")
cohens_d <- (mean_d - mean_b) / sd_pooled  # Flips the direction
print(paste("Cohen's d:", round(cohens_d, 4)))

# Convert sample sizes to numeric to avoid integer overflow
n1 <- as.numeric(n1)  # Before pandemic
n2 <- as.numeric(n2)  # During pandemic

# Standard error of Cohen's d
se_d <- sqrt((n1 + n2) / (n1 * n2) + (cohens_d^2) / (2 * (n1 + n2)))

# Calculate CIs
ci_99_lower_d <- cohens_d - z_99 * se_d
ci_99_upper_d <- cohens_d + z_99 * se_d

ci_95_lower_d <- cohens_d - z_95 * se_d  
ci_95_upper_d <- cohens_d + z_95 * se_d  

# Print results
cat("99% CI for Cohen's d:", ci_99_lower_d, "to", ci_99_upper_d, "\n")
cat("95% CI for Cohen's d:", ci_95_lower_d, "to", ci_95_upper_d, "\n") 

# Note: Confidence intervals for Cohen's d can be calculated using bootstrapping for more precise estimates, but this script uses an analytical approximation based on the standard error of d and Z-scores (valid for sufficiently large sample sizes and normal effect size distribution).

```

#### 2.1.3 Effect size (Cohen's d) - Vitamin D deficiency rates

```{r}
# Ensure sample sizes are numeric to avoid integer overflow
n1_def <- as.numeric(sum(df$period == "B"))
n2_def <- as.numeric(sum(df$period == "D"))

# Compute means and standard deviations for deficiency rates
mean_b_def <- mean(df$deficient[df$period == "B"])
mean_d_def <- mean(df$deficient[df$period == "D"])
sd_b_def <- sd(df$deficient[df$period == "B"])
sd_d_def <- sd(df$deficient[df$period == "D"])

# Pooled SD for deficiency
sd_pooled_def <- sqrt(((n1_def - 1) * sd_b_def^2 + (n2_def - 1) * sd_d_def^2) / (n1_def + n2_def - 2))

# Cohen's d (flipped sign for "During - Before")
cohens_d_def <- (mean_d_def - mean_b_def) / sd_pooled_def
print(paste("Cohen's d for deficiency:", round(cohens_d_def, 4)))

# Standard error for Cohen's d (now using numeric sample sizes)
se_d_def <- sqrt((n1_def + n2_def) / (n1_def * n2_def) + (cohens_d_def^2) / (2 * (n1_def + n2_def)))

# Confidence intervals for Cohen's d
ci_99_lower_d_def <- cohens_d_def - z_99 * se_d_def
ci_99_upper_d_def <- cohens_d_def + z_99 * se_d_def
ci_95_lower_d_def <- cohens_d_def - z_95 * se_d_def
ci_95_upper_d_def <- cohens_d_def + z_95 * se_d_def

# Print results
cat("99% CI for Cohen's d:", ci_99_lower_d_def, "to", ci_99_upper_d_def, "\n")
cat("95% CI for Cohen's d:", ci_95_lower_d_def, "to", ci_95_upper_d_def, "\n")

```


### 2.2 Mean levels stratified by age, gender and season
```{r}
# Compare mean vitamin D levels between periods using t-test
vitamin_d3_levels_B <- df %>% filter(period == "B") %>% pull(value_result)
vitamin_d3_levels_D <- df %>% filter(period == "D") %>% pull(value_result)

t_test_results <- t.test(vitamin_d3_levels_B, vitamin_d3_levels_D, na.rm = TRUE)
mean_comparison_p_value <- t_test_results$p.value

print(mean_comparison_p_value)


# Construct a contingency table for deficiencies in periods B and D
deficiency_table <- df %>% 
  group_by(period) %>% 
  summarise(deficient = sum(deficient), not_deficient = n() - sum(deficient)) %>%
  dplyr::select(-period) %>% 
  as.matrix()

chi_square_test_result <- chisq.test(deficiency_table)
  deficiency_comparison_p_value <- chi_square_test_result$p.value
  

print(deficiency_comparison_p_value)

# Mean Vitamin D levels by age group
vitamin_d3_age_means <- df %>%
  group_by(period, age_group) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

age_group_stats <- df %>%
  group_by(period, age_group) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )


# Mean Vitamin D Levels by gender group
vitamin_d3_gender_means <- df %>%
  group_by(period, gender) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

gender_group_stats <- df %>%
  filter(gender %in% c("M", "F")) %>%
  group_by(period, gender) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

# Mean Vitamin D Levels by season 
vitamin_d3_season_means <- df %>%
  group_by(period, season) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

season_group_stats <- df %>%
  group_by(period, season) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

print(age_group_stats)
print(gender_group_stats)
print(season_group_stats)

```

### 2.4 Statistical testing for age, gender and season
```{r}
# Statistical Testing within each sub-group

# Function to perform a t-test and return p-value
perform_t_test <- function(data_group_1, data_group_2) {
  test_result <- t.test(data_group_1, data_group_2, na.rm = TRUE)
  return(test_result$p.value)
}

# Calculate p-values for age groups
p_values_age_groups <- df %>%
  group_by(age_group) %>%
  summarise(
    t_test_p_value = perform_t_test(
      value_result[period == "B"],
      value_result[period == "D"]
    ),
    .groups = 'drop'  
  )

# Calculate p-values for gender groups
p_values_gender_groups <- df %>%
  filter(gender %in% c("M", "F")) %>%
  group_by(gender) %>%
  summarise(
    t_test_p_value = perform_t_test(
      value_result[period == "B"],
      value_result[period == "D"]
    ),
    .groups = 'drop'  
  )

# Calculate p-values for gender groups
p_values_season_groups <- df %>%
  group_by(season) %>%
  summarise(
    t_test_p_value = t.test(
      value_result[period == "B"],
      value_result[period == "D"],
    )$p.value,
    .groups = 'drop'
  )

# Print p-values
print(p_values_age_groups)
print(p_values_gender_groups)
print(p_values_season_groups)

# Exact p values

# LaTeX-ready scientific formatter: "p = 2.34 \\times 10^{-47}"
format_p_latex <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("$p = \\text{NA}$")
  if (p == 0) return("$p < 1 \times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)   # e.g. "4.91e-118"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- as.integer(parts[2])
  paste0("$p = ", base, " \times 10^{", exp, "}$")
}


# Helper: Welch t-test p-value (B vs D) for each level of a categorical var
welch_p_by_group <- function(data, group_var, value_var = "value_result",
                             period_var = "period") {
  stopifnot(all(c(group_var, value_var, period_var) %in% names(data)))
  d <- dplyr::filter(data, .data[[period_var]] %in% c("B","D"),
                     !is.na(.data[[group_var]]), !is.na(.data[[value_var]]))

  levs <- if (is.factor(d[[group_var]])) levels(d[[group_var]]) else sort(unique(d[[group_var]]))

  purrr::map_dfr(levs, function(lvl) {
    sub <- d %>% dplyr::filter(.data[[group_var]] == lvl)
    xB <- sub %>% dplyr::filter(.data[[period_var]] == "B") %>% dplyr::pull(.data[[value_var]])
    xD <- sub %>% dplyr::filter(.data[[period_var]] == "D") %>% dplyr::pull(.data[[value_var]])
    p  <- t.test(xB, xD, var.equal = FALSE)$p.value
    tibble::tibble(
      variable = group_var,
      level = as.character(lvl),
      p_value = p,
      p_formatted_latex = format_p_latex(p)
    )
  })
}

# ----- TOTAL (B vs D) -----
p_total <- t.test(value_result ~ period, data = df, var.equal = FALSE)$p.value
p_total_fmt <- format_p_latex(p_total)

# ----- BY AGE GROUP -----
p_age_tbl <- welch_p_by_group(df, "age_group")
# Optional: ensure your preferred order
p_age_tbl <- p_age_tbl %>%
  dplyr::mutate(level = factor(level, levels = c("[18, 39]", "[40, 59]", "60+"))) %>%
  dplyr::arrange(level)

# ----- BY GENDER (restrict to F/M just in case) -----
p_gender_tbl <- df %>%
  dplyr::filter(gender %in% c("F","M")) %>%
  welch_p_by_group("gender") %>%
  dplyr::mutate(level = dplyr::recode(level, "F" = "Female", "M" = "Male")) %>%
  dplyr::mutate(level = factor(level, levels = c("Female","Male"))) %>%
  dplyr::arrange(level)

# ----- BY SEASON -----
p_season_tbl <- welch_p_by_group(df, "season") %>%
  dplyr::mutate(level = factor(level, levels = c("Winter","Spring","Summer","Fall"))) %>%
  dplyr::arrange(level)

# Peek
p_total_fmt
p_age_tbl
p_gender_tbl
p_season_tbl

cat(
  "\nTOTAL mean comparison:\n  ", p_total_fmt, "\n",
  "\nBy age:\n  ",
  paste0(sprintf("%s: %s", as.character(p_age_tbl$level), p_age_tbl$p_formatted_latex), collapse = "\n  "),
  "\n\nBy gender:\n  ",
  paste0(sprintf("%s: %s", as.character(p_gender_tbl$level), p_gender_tbl$p_formatted_latex), collapse = "\n  "),
  "\n\nBy season:\n  ",
  paste0(sprintf("%s: %s", as.character(p_season_tbl$level), p_season_tbl$p_formatted_latex), collapse = "\n  "),
  "\n"
)

```

### 2.4 Median levels and interquartile ranges
```{r}
# Median and IQR for the full dataset, stratified by period
overall_median_iqr <- df %>%
  group_by(period) %>%
  summarise(
    median_value = median(value_result, na.rm = TRUE),
    Q1 = quantile(value_result, 0.25, na.rm = TRUE),
    Q3 = quantile(value_result, 0.75, na.rm = TRUE),
    .groups = 'drop'
  )

print(overall_median_iqr)

# Median and IQR by age group
age_group_median_iqr <- df %>%
  group_by(period, age_group) %>%
  summarise(
    median_value = median(value_result, na.rm = TRUE),
    Q1 = quantile(value_result, 0.25, na.rm = TRUE),
    Q3 = quantile(value_result, 0.75, na.rm = TRUE),
    .groups = 'drop'
  )

# Median and IQR by gender
gender_group_median_iqr <- df %>%
  filter(gender %in% c("M", "F")) %>%
  group_by(period, gender) %>%
  summarise(
    median_value = median(value_result, na.rm = TRUE),
    Q1 = quantile(value_result, 0.25, na.rm = TRUE),
    Q3 = quantile(value_result, 0.75, na.rm = TRUE),
    .groups = 'drop'
  )

# Median and IQR by season
season_group_median_iqr <- df %>%
  group_by(period, season) %>%
  summarise(
    median_value = median(value_result, na.rm = TRUE),
    Q1 = quantile(value_result, 0.25, na.rm = TRUE),
    Q3 = quantile(value_result, 0.75, na.rm = TRUE),
    .groups = 'drop'
  )

# Print all
print(age_group_median_iqr)
print(gender_group_median_iqr)
print(season_group_median_iqr)

# Overall Wilcoxon test
overall_wilcox <- wilcox.test(value_result ~ period, data = df)
print(overall_wilcox)

# Wilcoxon test by age group
age_group_wilcox <- df %>%
  filter(!is.na(age_group)) %>%
  group_by(age_group) %>%
  summarise(
    p_value = tryCatch(
      wilcox.test(value_result ~ period)$p.value,
      error = function(e) NA
    ),
    .groups = "drop"
  )
print(age_group_wilcox)

# Wilcoxon test by gender
gender_group_wilcox <- df %>%
  filter(gender %in% c("M", "F")) %>%
  group_by(gender) %>%
  summarise(
    p_value = tryCatch(
      wilcox.test(value_result ~ period)$p.value,
      error = function(e) NA
    ),
    .groups = "drop"
  )
print(gender_group_wilcox)

# Wilcoxon test by season
season_group_wilcox <- df %>%
  filter(!is.na(season)) %>%
  group_by(season) %>%
  summarise(
    p_value = tryCatch(
      wilcox.test(value_result ~ period)$p.value,
      error = function(e) NA
    ),
    .groups = "drop"
  )
print(season_group_wilcox)

```
```{r}
# Exact p values

format_p_latex <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("$p = \\text{NA}$")
  if (p == 0) return("$p < 1 \times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)  # e.g. "2.34e-47"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- as.integer(parts[2])
  paste0("$p = ", base, " \times 10^{", exp, "}$")
}

# Helper: Wilcoxon p-value (B vs D) per level of a categorical variable
wilcox_p_by_group <- function(data, group_var,
                              value_var = "value_result",
                              period_var = "period") {
  # iterate levels explicitly for robustness
  levs <- data[[group_var]] %>% unique() %>% sort()
  map_dfr(levs, function(lvl) {
    sub <- data %>%
      filter(!is.na(.data[[group_var]]),
             .data[[group_var]] == lvl,
             .data[[period_var]] %in% c("B","D"),
             !is.na(.data[[value_var]]))
    # need both periods present
    have_B <- any(sub[[period_var]] == "B")
    have_D <- any(sub[[period_var]] == "D")
    p <- if (!have_B || !have_D) {
      NA_real_
    } else {
      # exact=FALSE avoids warnings for large n / ties
      wilcox.test(as.formula(paste(value_var, "~", period_var)),
                  data = sub, exact = FALSE)$p.value
    }
    tibble(
      variable = group_var,
      level = as.character(lvl),
      p_value = p,
      p_formatted_latex = format_p_latex(p)
    )
  }) %>% arrange(level)
}

# ----- TOTAL (B vs D) -----
p_total_wilcox <- wilcox.test(value_result ~ period, data = df, exact = FALSE)$p.value
p_total_wilcox_fmt <- format_p_latex(p_total_wilcox)

# ----- BY AGE -----
p_age_wilcox <- wilcox_p_by_group(df, "age_group") %>%
  mutate(level = factor(level, levels = c("[18, 39]", "[40, 59]", "60+"))) %>%
  arrange(level)

# ----- BY GENDER (restrict to F/M, then relabel) -----
p_gender_wilcox <- df %>%
  filter(gender %in% c("F","M")) %>%
  wilcox_p_by_group("gender") %>%
  mutate(level = recode(level, "F" = "Female", "M" = "Male"),
         level = factor(level, levels = c("Female","Male"))) %>%
  arrange(level)

# ----- BY SEASON -----
p_season_wilcox <- wilcox_p_by_group(df, "season") %>%
  mutate(level = factor(level, levels = c("Winter","Spring","Summer","Fall"))) %>%
  arrange(level)


p_total_wilcox_fmt
p_age_wilcox
p_gender_wilcox
p_season_wilcox

# ----- COPY/PASTE HELP: emit LaTeX-ready strings for your table -----
cat(
  "\nWilcoxon p-values for median/IQR table:\n",
  "Total: ", p_total_wilcox_fmt, "\n\n",
  "By age:\n  ",
  paste0(sprintf("%s: %s", as.character(p_age_wilcox$level), p_age_wilcox$p_formatted_latex), collapse = "\n  "),
  "\n\nBy gender:\n  ",
  paste0(sprintf("%s: %s", as.character(p_gender_wilcox$level), p_gender_wilcox$p_formatted_latex), collapse = "\n  "),
  "\n\nBy season:\n  ",
  paste0(sprintf("%s: %s", as.character(p_season_wilcox$level), p_season_wilcox$p_formatted_latex), collapse = "\n  "),
  "\n"
)

```


### 2.5 Deficiency rates for gender, age and season groups
```{r}
# Calculate the deficiency rates for each gender group
deficiency_gender_groups <- df %>%
  group_by(gender, period) %>%
  summarise(deficient_count = sum(deficient), total_count = n(), .groups = 'drop') %>%
  pivot_wider(
    names_from = period,
    values_from = c(deficient_count, total_count),
    values_fill = list(deficient_count = 0, total_count = 0)
  ) %>%
  mutate(
    rate_B = if_else(total_count_B > 0, deficient_count_B / total_count_B * 100, 0),
    rate_D = if_else(total_count_D > 0, deficient_count_D / total_count_D * 100, 0)
  ) %>%
  # Calculate p-values
  rowwise() %>%
  mutate(p_value = chisq.test(rbind(c(deficient_count_B, total_count_B - deficient_count_B), 
                                    c(deficient_count_D, total_count_D - deficient_count_D)))$p.value) %>%
  ungroup()

print(deficiency_gender_groups)
```

```{r}
# Calculate the deficiency rates for each age group
deficiency_age_groups <- df %>%
  group_by(age_group, period) %>%
  summarise(deficient_count = sum(deficient), total_count = n(), .groups = 'drop') %>%
  pivot_wider(
    names_from = period,
    values_from = c(deficient_count, total_count),
    values_fill = list(deficient_count = 0, total_count = 0) 
  ) %>%
  mutate(
    rate_B = if_else(total_count_B > 0, deficient_count_B / total_count_B * 100, 0),
    rate_D = if_else(total_count_D > 0, deficient_count_D / total_count_D * 100, 0)
  ) %>%
  rowwise() %>%
  mutate(p_value = chisq.test(rbind(c(deficient_count_B, total_count_B - deficient_count_B), 
                                    c(deficient_count_D, total_count_D - deficient_count_D)))$p.value) %>%
  ungroup()

print(deficiency_age_groups)
```


```{r}
# Calculate the deficiency rates for each season group
deficiency_season_groups <- df %>%
  group_by(season, period) %>%
  summarise(deficient_count = sum(deficient), total_count = n(), .groups = 'drop') %>%
  pivot_wider(
    names_from = period,
    values_from = c(deficient_count, total_count),
    values_fill = list(deficient_count = 0, total_count = 0)
  ) %>%
  mutate(
    rate_B = if_else(total_count_B > 0, deficient_count_B / total_count_B * 100, 0),
    rate_D = if_else(total_count_D > 0, deficient_count_D / total_count_D * 100, 0)
  ) %>%
  # Calculate p-values
  rowwise() %>%
  mutate(p_value = chisq.test(rbind(c(deficient_count_B, total_count_B - deficient_count_B), 
                                    c(deficient_count_D, total_count_D - deficient_count_D)))$p.value) %>%
  ungroup()

print(deficiency_season_groups)
```

```{r}
# Exact p values

# LaTeX-ready scientific formatter (single backslash)
format_p_latex <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("$p = \\text{NA}$")
  if (p == 0) return("$p < 1 \times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)   # e.g. "4.91e-118"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- as.integer(parts[2])
  paste0("$p = ", base, " \times 10^{", exp, "}$")
}

# Deficiency rates + chi-square p (B vs D) for each level of a grouping var
deficiency_rates_by_group <- function(data, group_var,
                                      period_var = "period",
                                      deficient_var = "deficient") {
  gsym <- sym(group_var)
  psym <- sym(period_var)
  dsym <- sym(deficient_var)

  wide <- data %>%
    group_by(!!gsym, !!psym) %>%
    summarise(
      deficient_count = sum(!!dsym, na.rm = TRUE),
      total_count     = n(),
      .groups = "drop"
    ) %>%
    tidyr::pivot_wider(
      names_from  = !!psym,                         # <- tidy-eval (no .data)
      values_from = c(deficient_count, total_count),
      values_fill = list(deficient_count = 0L, total_count = 0L)
    ) %>%
    mutate(
      rate_B = if_else(total_count_B > 0, 100 * deficient_count_B / total_count_B, 0),
      rate_D = if_else(total_count_D > 0, 100 * deficient_count_D / total_count_D, 0)
    )

  out <- wide %>%
    rowwise() %>%
    mutate(
      p_value = {
        # guard against missing period columns
        if (!all(c("deficient_count_B","total_count_B",
                   "deficient_count_D","total_count_D") %in% names(cur_data()))) {
          NA_real_
        } else if (total_count_B == 0 || total_count_D == 0) {
          NA_real_
        } else {
          suppressWarnings(
            chisq.test(
              rbind(
                c(deficient_count_B, total_count_B - deficient_count_B),
                c(deficient_count_D, total_count_D - deficient_count_D)
              ),
              correct = FALSE
            )$p.value
          )
        }
      }
    ) %>%
    ungroup() %>%
    mutate(
      p_formatted_latex = map_chr(p_value, format_p_latex)  # <- row-wise formatting
    ) %>%
    rename(level = !!gsym)

  out
}

# ---------- TOTAL (B vs D) ----------
def_tab_total <- df %>%
  count(period, deficient, name = "n") %>%
  tidyr::pivot_wider(names_from = period, values_from = n, values_fill = 0L) %>%
  mutate(
    B_def = `B`[deficient == 1], B_ok = `B`[deficient == 0],
    D_def = `D`[deficient == 1], D_ok = `D`[deficient == 0]
  ) %>%
  summarise(
    B_def = first(B_def) %||% 0L, B_ok = first(B_ok) %||% 0L,
    D_def = first(D_def) %||% 0L, D_ok = first(D_ok) %||% 0L
  )

def_mat_total <- matrix(c(def_tab_total$B_def, def_tab_total$B_ok,
                          def_tab_total$D_def, def_tab_total$D_ok),
                        nrow = 2, byrow = TRUE)

p_def_total      <- suppressWarnings(chisq.test(def_mat_total, correct = FALSE)$p.value)
p_def_total_fmt  <- format_p_latex(p_def_total)

# ---------- BY GROUP ----------
# Age
deficiency_age_groups <- deficiency_rates_by_group(df, "age_group") %>%
  mutate(level = factor(level, levels = c("[18, 39]", "[40, 59]", "60+"))) %>%
  arrange(level)

# Gender (F/M only, then relabel)
deficiency_gender_groups <- df %>%
  filter(gender %in% c("F","M")) %>%
  deficiency_rates_by_group("gender") %>%
  mutate(level = recode(level, "F" = "Female", "M" = "Male"),
         level = factor(level, levels = c("Female","Male"))) %>%
  arrange(level)

# Season
deficiency_season_groups <- deficiency_rates_by_group(df, "season") %>%
  mutate(level = factor(level, levels = c("Winter","Spring","Summer","Fall"))) %>%
  arrange(level)

# Quick peek (optional)
p_def_total_fmt
deficiency_age_groups
deficiency_gender_groups
deficiency_season_groups


```


### 2.6 Plots: Vit D levels over time
```{r}
# Vit D Levels Over Time (stratified by age and gender) - with Standard Error

# Calculate mean and standard error for Vitamin D levels by gender and season-year
gender_stats <- df %>%
  group_by(season_year, gender) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    se_value = sd(value_result, na.rm = TRUE) / sqrt(n()),  # Standard error calculation
    .groups = 'drop'
  )

# Plot Vitamin D levels over time by gender group (with standard error)
p_gender <- ggplot(gender_stats, aes(x = season_year, y = mean_value, color = gender, fill = gender, group = gender)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) +  
  geom_ribbon(aes(ymin = mean_value - se_value, ymax = mean_value + se_value), alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)"
  ) +
  ylim(0, 60) +  
  theme_minimal() +
  scale_fill_manual(values = c("M" = "blue", "F" = "red")) +  
  scale_color_manual(values = c("M" = "blue", "F" = "red")) +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

# Calculate mean and standard error for Vitamin D levels by age group and season-year
age_group_stats <- df %>%
  group_by(season_year, age_group) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    se_value = sd(value_result, na.rm = TRUE) / sqrt(n()),  # Standard error calculation
    .groups = 'drop'
  )

# Plot Vitamin D levels over time by age group (with standard error)
p_age <- ggplot(age_group_stats, aes(x = season_year, y = mean_value, color = age_group, fill = age_group, group = age_group)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) +  
  geom_ribbon(aes(ymin = mean_value - se_value, ymax = mean_value + se_value), alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)"
  ) +
  ylim(0, 60) + 
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +  
  scale_color_brewer(palette = "Set2") +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
# Vit D Levels Over Time (stratified by age and gender) - with Standard Deviation

# Calculate mean and standard deviation for Vitamin D levels by gender and season-year
gender_stats <- df %>%
  group_by(season_year, gender) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE), 
    .groups = 'drop'
  )

# Plot Vitamin D levels over time by gender group (with standard deviation)
ggplot(gender_stats, aes(x = season_year, y = mean_value, color = gender, fill = gender, group = gender)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) + 
  geom_ribbon(aes(ymin = mean_value - sd_value, ymax = mean_value + sd_value), alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)"
  ) +
  ylim(0, 60) +  
  theme_minimal() +
  scale_fill_manual(values = c("M" = "blue", "F" = "red")) +  
  scale_color_manual(values = c("M" = "blue", "F" = "red")) +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

# Calculate mean and standard deviation for Vitamin D levels by age group and season-year
age_group_stats <- df %>%
  group_by(season_year, age_group) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    sd_value = sd(value_result, na.rm = TRUE),  # Standard deviation calculation
    .groups = 'drop'
  )

# Plot Vitamin D levels over time by age group (with standard deviation)
ggplot(age_group_stats, aes(x = season_year, y = mean_value, color = age_group, fill = age_group, group = age_group)) +
  geom_line(linewidth = 1) +  
  geom_point(size = 2) +  
  geom_ribbon(aes(ymin = mean_value - sd_value, ymax = mean_value + sd_value), alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)"
  ) +
  ylim(0, 60) +  
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +  
  scale_color_brewer(palette = "Set2") +
  theme(legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))

```

#### FIGURE S1
```{r}
# Vit D Levels Over Time (not stratified by age and gender)

# Calculate mean and standard error for Vitamin D levels for the entire cohort by season-year
cohort_stats <- df %>%
  group_by(season_year) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    se_value = sd(value_result, na.rm = TRUE) / sqrt(n()),
    sd_value = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

print(cohort_stats)

# Plot Vitamin D levels over time for the entire cohort with standard error ribbon
ggplot(cohort_stats, aes(x = season_year, y = mean_value, group = 1)) +
  geom_line(color = "#1f77b4", linewidth = 1) +  
  geom_point(color = "#ff7f0e", size = 2) +  
  geom_ribbon(aes(ymin = mean_value - se_value, ymax = mean_value + se_value), fill = "#ff7f0e", alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)",
  ) +
  ylim(0, 60) + 
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )


# Plot Vitamin D levels over time for the entire cohort with standard deviation ribbon
ggplot(cohort_stats, aes(x = season_year, y = mean_value, group = 1)) +
  geom_line(color = "#1f77b4", linewidth = 1) +  
  geom_point(color = "#ff7f0e", size = 2) +  
  geom_ribbon(aes(ymin = mean_value - sd_value, ymax = mean_value + sd_value), fill = "#ff7f0e", alpha = 0.2) +  
  labs(
    x = "Season",
    y = "Vitamin D Level (µg/l)",
  ) +
  ylim(0, 60) +  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )


```

### 2.7 FIGURE S2: Plots: Early vs. late pandemic

```{r}
# Define early and late pandemic periods
early_period <- df %>%
  filter(test_date >= as.Date("2020-03-01") & test_date <= as.Date("2020-08-31")) %>%
  mutate(pandemic_phase = "Early Pandemic")

late_period <- df %>%
  filter(test_date >= as.Date("2021-09-01") & test_date <= as.Date("2022-02-28")) %>%
  mutate(pandemic_phase = "Late Pandemic")

# Combine into one dataset
pandemic_comparison_df <- bind_rows(early_period, late_period)

# Summary statistics
pandemic_summary <- pandemic_comparison_df %>%
  group_by(pandemic_phase) %>%
  summarise(
    n = n(),
    mean_vitamin_d = mean(value_result, na.rm = TRUE),
    sd_vitamin_d = sd(value_result, na.rm = TRUE),
    median_vitamin_d = median(value_result, na.rm = TRUE),
    iqr_vitamin_d = IQR(value_result, na.rm = TRUE)
  )

print(pandemic_summary)

# Statistical testing
t_test_pandemic <- t.test(value_result ~ pandemic_phase, data = pandemic_comparison_df)
wilcox_test_pandemic <- wilcox.test(value_result ~ pandemic_phase, data = pandemic_comparison_df, exact = FALSE)

cat("T-test p-value:", t_test_pandemic$p.value, "\n")
cat("Wilcoxon test p-value:", wilcox_test_pandemic$p.value, "\n")

# Define matching colors
pandemic_colors <- c("Early Pandemic" = "#1f77b4", "Late Pandemic" = "#ff7f0e")

# Determine significance stars based on Wilcoxon test
signif_star <- if (wilcox_test_pandemic$p.value < 0.001) {
  "***"
} else if (wilcox_test_pandemic$p.value < 0.01) {
  "**"
} else if (wilcox_test_pandemic$p.value < 0.05) {
  "*"
} else {
  "n.s."
}

# Boxplot with y-axis cropped and significance stars
p_pandemic_box <- ggplot(pandemic_comparison_df, aes(x = pandemic_phase, y = value_result, fill = pandemic_phase)) +
  geom_boxplot(width = 0.5, alpha = 0.8, outlier.shape = NA) +
  ylim(NA, 75) +  # Crop y-axis at 75
  
  # Add significance annotation
  annotate("text", x = 1.5, y = 73, label = signif_star, size = 6, fontface = "bold") +
  annotate("segment", x = 1, xend = 2, y = 70, yend = 70) +  # horizontal line
  annotate("segment", x = 1, xend = 1, y = 70, yend = 68) +  # left vertical
  annotate("segment", x = 2, xend = 2, y = 70, yend = 68) +  # right vertical
  
  # Labels and colors
  labs(
    x = "Pandemic Phase",
    y = "Vitamin D Level (µg/l)",
    fill = "Pandemic Phase"
  ) +
  scale_fill_manual(values = pandemic_colors, aesthetics = c("fill", "color")) +
  
  # Theme settings
  theme_classic() +
  theme(
    text = element_text(size = 10),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_line(color = "grey90", size = 0.2),
    panel.grid.minor.y = element_line(color = "grey95", size = 0.1),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  )

p_pandemic_box
```


### 2.8 FIGURE 3: Comparison Vit D levels before and during the pandemic

```{r}
# helper to format p in scientific notation with superscript exponent
format_p <- function(p, digits = 2) {
  sup <- c("0"="⁰","1"="¹","2"="²","3"="³","4"="⁴","5"="⁵",
           "6"="⁶","7"="⁷","8"="⁸","9"="⁹","-"="⁻","+"="⁺")
  to_sup <- function(n) {
    s <- strsplit(as.character(n), "")[[1]]
    paste0(sup[s], collapse = "")
  }
  fmt_one <- function(x) {
    if (is.na(x)) return(NA_character_)
    s <- formatC(x, format = "e", digits = digits)     # e.g., "5.60e-18"
    parts <- strsplit(s, "e")[[1]]
    mant  <- parts[1]
    expo  <- as.integer(parts[2])
    paste0("p = ", mant, " × 10", to_sup(expo))
  }
  vapply(p, fmt_one, character(1))
}

# --- data prep (same filtering as before) ------------------------------------
df_filtered <- df %>%
  dplyr::filter(value_result >= 0, value_result <= 70) %>%
  dplyr::mutate(gender_age_group = paste(gender, age_group, sep = " - "))

# Welch two-sided t-test per subgroup (B vs D)
t_test_gender_age_group <- df_filtered %>%
  dplyr::group_by(gender_age_group) %>%
  dplyr::summarise(
    p_value = t.test(value_result[period == "B"],
                     value_result[period == "D"],
                     na.rm = TRUE)$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(p_label = format_p(p_value))

# mean changes for facet captions
gender_age_group_changes <- df_filtered %>%
  dplyr::group_by(gender_age_group, period) %>%
  dplyr::summarise(mean_value = mean(value_result, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = period, values_from = mean_value) %>%
  dplyr::rename(mean_B = B, mean_D = D) %>%
  dplyr::mutate(
    absolute_change  = mean_D - mean_B,
    percentage_change = (mean_D - mean_B) / mean_B * 100
  )

# merge in p labels and build 3-line facet labels
facet_info <- dplyr::left_join(gender_age_group_changes, t_test_gender_age_group,
                               by = "gender_age_group")

facet_labels <- setNames(
  paste0(
    facet_info$gender_age_group,
    "\nAbs. Change: ", round(facet_info$absolute_change, 2), " µg/l;\n ",
    "Rel. Change: ", round(facet_info$percentage_change, 1), "%\n",
    facet_info$p_label
  ),
  facet_info$gender_age_group
)

# --- plot (unchanged geometry; p-values now live in the strip label) ---------
p_gender_age_violin_box <- ggplot(df_filtered, aes(x = period, y = value_result, fill = period)) +
  geom_violin(position = position_dodge(width = 0.8), width = 0.6, trim = FALSE) +
  geom_boxplot(position = position_dodge(width = 0.8), width = 0.15, alpha = 0.5, outlier.shape = NA) +
  stat_summary(fun.data = "mean_sdl", geom = "pointrange", position = position_dodge(width = 0.8)) +
  labs(x = "Period", y = "Vitamin D Level (µg/l)") +
  theme_minimal() +
  scale_fill_manual(values = c("B" = "#1f77b4", "D" = "#ff7f0e"),
                    name = "Period",
                    labels = c("B" = "Pre-Pandemic", "D" = "Pandemic")) +
  theme(
    legend.title = element_blank(),
    strip.text = element_text(size = 7),     # same style for all three lines in strip
    axis.text.x = element_blank(),
    axis.ticks.length = unit(0.2, "cm")
  ) +
  facet_wrap(~ gender_age_group, scales = "free", labeller = as_labeller(facet_labels))

print(p_gender_age_violin_box)

```


### 2.9 FIGURE S3: Plots: Seasonality 

```{r}
# Filter data to remove outliers (values below 0 or above 70)
df_filtered <- df %>%
  filter(value_result >= 0 & value_result <= 70)

# Prepare the data for the updated plot
season_year_stats <- df_filtered %>%
  filter(!(cohort_year == 2018 & season == "Winter")) %>% # Exclude Winter 2018
  mutate(
    cohort_year = factor(cohort_year),  
    period_label = case_when(
      period == "B" ~ "Pre-Pandemic",
      period == "D" ~ "Pandemic",
      TRUE ~ NA_character_
    ),
    period_label = factor(period_label, levels = c("Pre-Pandemic", "Pandemic"))
  )

# Calculate mean and median differences between pre-pandemic and pandemic periods for each season
seasonal_changes <- season_year_stats %>%
  group_by(season, period_label) %>%
  summarise(
    mean_value = mean(value_result, na.rm = TRUE),
    median_value = median(value_result, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = period_label, 
    values_from = c(mean_value, median_value)
  ) %>%
  rename(
    mean_value_Pre_Pandemic = `mean_value_Pre-Pandemic`,
    mean_value_Pandemic = `mean_value_Pandemic`,
    median_value_Pre_Pandemic = `median_value_Pre-Pandemic`,
    median_value_Pandemic = `median_value_Pandemic`
  ) %>%
  mutate(
    mean_diff = mean_value_Pandemic - mean_value_Pre_Pandemic,
    median_diff = median_value_Pandemic - median_value_Pre_Pandemic
  )

seasonal_changes <- seasonal_changes %>%
  dplyr::select(season, mean_diff) %>%
  mutate(mean_diff = round(mean_diff, 2))

# Create facet labels with mean difference included
facet_labels <- setNames(
  paste0(c("Winter", "Spring", "Summer", "Fall"), "\nMean Diff: ", seasonal_changes$mean_diff),
  c("Winter", "Spring", "Summer", "Fall")
)

# Updated plot with mean difference in facet labels
p_seasonal_violin_box <- ggplot(season_year_stats, aes(x = cohort_year, y = value_result, fill = period_label)) +
  geom_violin(alpha = 0.8, trim = TRUE, scale = "width", position = position_dodge(width = 0.8)) +
  geom_boxplot(width = 0.2, position = position_dodge(width = 0.8), alpha = 0.6, outlier.shape = NA) + 
  facet_wrap(~ season, ncol = 2, labeller = as_labeller(facet_labels)) +
  geom_vline(
  xintercept = seq_along(unique(season_year_stats$cohort_year)) - 0.5,
  color = "grey85",
  linetype = "dashed",
  size = 0.3
) +
  
  # Labels and colors
  labs(
    x = "Year",
    y = "Vitamin D Level (µg/l)",
    fill = "Period"
  ) +
  scale_fill_manual(
  values = c("Pre-Pandemic" = "#1f77b4", "Pandemic" = "#ff7f0e"),
  aesthetics = c("fill", "color")
) +
  
  # Adjust theme
  theme_classic() +
  theme(
    text = element_text(size = 10),
    legend.text = element_text(size = 8),
    panel.grid.major.y = element_line(color = "grey90", size = 0.2),
    panel.grid.minor.y = element_line(color = "grey95", size = 0.1),
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold")
  )

print(p_seasonal_violin_box)

```

## 3. Propensity score matching (PSM)

### 3.1 Matching
```{r}
# performing exact matching parameters on the entire sample
test_match_obj <- matchit(period ~ age_group + gender + test_month, data = df, method = "nearest", distance = "glm", exact = c("gender", "age_group", "test_month")) 

matched_data <- match.data(test_match_obj)

# Calculate the sample sizes
original_size <- nrow(df)  # Total original sample size
matched_size <- nrow(matched_data)  # Total matched sample size

# Calculate the number and percentage of observations lost
num_lost <- original_size - matched_size
percent_lost <- (num_lost / original_size) * 100

# Print results
cat("Original Sample Size: ", original_size, "\n")
cat("Matched Sample Size: ", matched_size, "\n")
cat("Number of Observations Lost: ", num_lost, "\n")
cat("Percentage Lost: ", percent_lost, "%\n")

# Summary
summary(test_match_obj)
```
#### 3.1.1 Statistical testing

```{r}
# Calculate summaries for Age with Chi-squared Test for Independence
age_summary <- matched_data %>%
  group_by(period, age_group) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
age_table <- xtabs(count ~ period + age_group, data = age_summary)
age_chi_sq_test <- tryCatch(chisq.test(age_table), error = function(e) e)
age_p_value <- ifelse(class(age_chi_sq_test) == "try-error", NA, age_chi_sq_test$p.value)

# Calculate summaries for Gender with Chi-squared Test
gender_summary <- matched_data %>%
  group_by(period, gender) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
gender_table <- xtabs(count ~ period + gender, data = gender_summary)
gender_chi_sq_test <- tryCatch(chisq.test(gender_table), error = function(e) e)
gender_p_value <- ifelse(class(gender_chi_sq_test) == "try-error", NA, gender_chi_sq_test$p.value)

# Calculate summaries for Test month with Chi-squared Test
test_month_summary <- matched_data %>%
  group_by(period, test_month) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
test_month_table <- xtabs(count ~ period + test_month, data = test_month_summary)
test_month_chi_sq_test <- tryCatch(chisq.test(test_month_table), error = function(e) e)
test_month_p_value <- ifelse(class(test_month_chi_sq_test) == "try-error", NA, test_month_chi_sq_test$p.value)

# Calculate summaries for Season with Chi-squared Test
season_summary <- matched_data %>%
  group_by(period, season) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
season_table <- xtabs(count ~ period + season, data = season_summary)
season_chi_sq_test <- tryCatch(chisq.test(season_table), error = function(e) e)
season_p_value <- ifelse(class(season_chi_sq_test) == "try-error", NA, season_chi_sq_test$p.value)

# Calculate mean and standard deviation for Vitamin D levels and perform t-test
vitamin_d_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    mean_vitamin_d = mean(value_result, na.rm = TRUE),
    sd_vitamin_d = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )
vitamin_d_t_test <- t.test(value_result ~ period, data = matched_data)
vitamin_d_p_value <- vitamin_d_t_test$p.value

# Calculate the percentage of deficiencies for each period and perform Chi-squared Test
deficiency_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    total = n(),
    deficient_count = sum(deficient == 1, na.rm = TRUE),
    non_deficient_count = sum(deficient == 0, na.rm = TRUE),
    deficiency_rate = deficient_count / total * 100,
    .groups = 'drop'
  )

# Create a contingency table for the Chi-squared test
deficiency_table <- xtabs(~ period + deficient, data = matched_data)

# Conduct Chi-squared Test and append the result
deficiency_chi_sq_test <- tryCatch(chisq.test(deficiency_table), error = function(e) e)
deficiency_p_value <- ifelse(class(deficiency_chi_sq_test) == "try-error", NA, deficiency_chi_sq_test$p.value)

# Print the summaries and p-values
print(age_summary)
print(paste("Age Group P-value:", age_p_value))
print(gender_summary)
print(paste("Gender P-value:", gender_p_value))
print(test_month_summary)
print(paste("Test Month P-value:", test_month_p_value))
print(season_summary)
print(paste("Season P-value:", season_p_value))
print(vitamin_d_summary)
print(paste("Vitamin D Levels P-value:", vitamin_d_p_value))
print(deficiency_summary)
print(paste("Deficiency Rate P-value:", deficiency_p_value))

```
```{r}
# Exact p values

# LaTeX-ready formatter
format_p_latex <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("$p = \\text{NA}$")
  if (p == 0) return("$p < 1 \times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)  # e.g. "2.34e-47"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- as.integer(parts[2])
  paste0("$p = ", base, " \times 10^{", exp, "}$")
}

# --- Vitamin D levels (Welch t-test B vs D) ---
vitamin_d_t_test <- tryCatch(
  t.test(value_result ~ period, data = matched_data, var.equal = FALSE),
  error = function(e) e
)
vitamin_d_p_value <- if (inherits(vitamin_d_t_test, "htest")) vitamin_d_t_test$p.value else NA_real_
vitamin_d_p_latex <- format_p_latex(vitamin_d_p_value)

# --- Deficiency rate (Pearson χ² on 2x2; no Yates correction) ---
def_table <- tryCatch(
  xtabs(~ period + deficient, data = matched_data),
  error = function(e) e
)

def_chi <- if (!inherits(def_table, "error")) {
  tryCatch(chisq.test(def_table, correct = FALSE), error = function(e) e)
} else def_table

deficiency_p_value <- if (inherits(def_chi, "htest")) def_chi$p.value else NA_real_
deficiency_p_latex <- format_p_latex(deficiency_p_value)

cat(
  "\nLaTeX-ready p-values (matched analysis):\n",
  "Vitamin D levels: ", vitamin_d_p_latex, "\n",
  "Deficiency rate:  ", deficiency_p_latex, "\n",
  sep = ""
)

```



#### 3.1.2 Effect size (SMD) - Mean vitamin D levels

```{r}
# Effect Size

# Calculate means and SDs for matched data
mean_matched_b <- mean(matched_data$value_result[matched_data$period == "B"])
mean_matched_d <- mean(matched_data$value_result[matched_data$period == "D"])
sd_matched_b <- sd(matched_data$value_result[matched_data$period == "B"])
sd_matched_d <- sd(matched_data$value_result[matched_data$period == "D"])

# Pooled SD for matched data
n1_matched <- nrow(matched_data[matched_data$period == "B", ])
n2_matched <- nrow(matched_data[matched_data$period == "D", ])
sd_matched_pooled <- sqrt(((n1_matched - 1) * sd_matched_b^2 + 
                           (n2_matched - 1) * sd_matched_d^2) / 
                          (n1_matched + n2_matched - 2))

# SMD (flipped sign for "During - Before")
smd <- (mean_matched_d - mean_matched_b) / sd_matched_pooled  # Flips the direction
print(paste("Standardized Mean Difference (SMD):", round(smd, 4)))

# Standard error for SMD (same formula as Cohen's d)
se_smd <- sqrt((n1 + n2) / (n1 * n2) + (smd^2) / (2 * (n1 + n2)))

# Confidence intervals
ci_99_lower_smd <- smd - z_99 * se_smd
ci_99_upper_smd <- smd + z_99 * se_smd

ci_95_lower_smd <- smd - z_95 * se_smd  
ci_95_upper_smd <- smd + z_95 * se_smd  

# Print results
cat("99% CI for SMD:", ci_99_lower_smd, "to", ci_99_upper_smd, "\n")
cat("95% CI for SMD:", ci_95_lower_smd, "to", ci_95_upper_smd, "\n")  

# Note: Confidence intervals for the Standardized Mean Difference (SMD) are calculated using the formula:CI = SMD ± Z * SE_SMD where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_SMD is the standard error of the SMD estimate.
```

#### 3.1.3 Effect size (SMD) - vitamin D deficiency rate

```{r}
# Ensure sample sizes are numeric to avoid integer overflow
n1_matched_def <- as.numeric(nrow(matched_data[matched_data$period == "B", ]))
n2_matched_def <- as.numeric(nrow(matched_data[matched_data$period == "D", ]))

# Compute means and standard deviations for matched deficiency rates
mean_matched_b_def <- mean(matched_data$deficient[matched_data$period == "B"])
mean_matched_d_def <- mean(matched_data$deficient[matched_data$period == "D"])
sd_matched_b_def <- sd(matched_data$deficient[matched_data$period == "B"])
sd_matched_d_def <- sd(matched_data$deficient[matched_data$period == "D"])

# Pooled SD for matched deficiency data
sd_matched_pooled_def <- sqrt(((n1_matched_def - 1) * sd_matched_b_def^2 + 
                               (n2_matched_def - 1) * sd_matched_d_def^2) / 
                              (n1_matched_def + n2_matched_def - 2))

# SMD for deficiency (flipped sign for "During - Before")
smd_def <- (mean_matched_d_def - mean_matched_b_def) / sd_matched_pooled_def
print(paste("SMD for deficiency:", round(smd_def, 4)))

# Standard error for SMD
se_smd_def <- sqrt((n1_matched_def + n2_matched_def) / (n1_matched_def * n2_matched_def) + (smd_def^2) / (2 * (n1_matched_def + n2_matched_def)))

# Confidence intervals for SMD
ci_99_lower_smd_def <- smd_def - z_99 * se_smd_def
ci_99_upper_smd_def <- smd_def + z_99 * se_smd_def
ci_95_lower_smd_def <- smd_def - z_95 * se_smd_def
ci_95_upper_smd_def <- smd_def + z_95 * se_smd_def

# Print results
cat("99% CI for SMD:", ci_99_lower_smd_def, "to", ci_99_upper_smd_def, "\n")
cat("95% CI for SMD:", ci_95_lower_smd_def, "to", ci_95_upper_smd_def, "\n")

# Note: Confidence intervals for the Standardized Mean Difference (SMD) are calculated using the formula:CI = SMD ± Z * SE_SMD where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_SMD is the standard error of the SMD estimate.
```


## 4.Causal Machine Learning: Random Forest ON UNMATCHED DATA

### 4.1 Change in mean vitamin D levels
```{r}
# Treatment indicator
df$treatment <- ifelse(df$period == "D", 1, 0)

# Create design matrix from categorical variables
X <- model.matrix(~ age_group + gender + season, data = df)[, -1]  # Remove intercept column

# Outcome
Y <- df$value_result

# Run causal forest
cf_model <- causal_forest(
  X = X,
  Y = Y,
  W = df$treatment,
  num.trees = 2000
)

# Average Treatment Effect (ATE)
ate_result <- average_treatment_effect(cf_model, target.sample = "all")
ate_estimate <- ate_result[1]
ate_std_err <- ate_result[2]

# Pre-pandemic average
average_value_result <- df %>%
  filter(period == "B") %>%
  summarise(avg_value = mean(value_result, na.rm = TRUE)) %>%
  pull(avg_value)

# Print results
cat("ATE Estimate:", round(ate_estimate, 4), "\n")
cat("Percentage change:", round((ate_estimate / average_value_result) * 100, 2), "%\n")

# Confidence intervals
z_95 <- 1.960
z_99 <- 2.576

ci_99_lower_ate <- ate_estimate - z_99 * ate_std_err
ci_99_upper_ate <- ate_estimate + z_99 * ate_std_err

ci_95_lower_ate <- ate_estimate - z_95 * ate_std_err
ci_95_upper_ate <- ate_estimate + z_95 * ate_std_err

cat("99% CI for ATE:", round(ci_99_lower_ate, 4), "to", round(ci_99_upper_ate, 4), "\n")
cat("95% CI for ATE:", round(ci_95_lower_ate, 4), "to", round(ci_95_upper_ate, 4), "\n")

# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.
```

#### 4.1.1 FIGURE 5b: Feature importance

```{r}
# Variable importance for each dummy variable
importance <- variable_importance(cf_model)

# Get column names from the design matrix
X_names <- colnames(X)

# Combine dummy variable importances by original variable
grouped_importance <- data.frame(variable = X_names, importance = importance) %>%
  mutate(group = case_when(
    grepl("^age_group", variable) ~ "Age Group",
    grepl("^gender", variable) ~ "Gender",
    grepl("^season", variable) ~ "Season"
  )) %>%
  group_by(group) %>%
  summarise(total_importance = sum(importance)) %>%
  ungroup() %>%
  mutate(percentage = (total_importance / sum(total_importance)) * 100)

# Plot aggregated feature importance
ggplot(grouped_importance, aes(x = reorder(group, total_importance), y = percentage, fill = group)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), hjust = -0.2, size = 5) +
  labs(
    x = "Covariate",
    y = "Feature importance (explained variance in %)"
  ) +
  scale_fill_manual(values = c("Age Group" = "#1f77b4", "Gender" = "#1f77b4", "Season" = "#1f77b4")) +
  coord_flip() +
  ylim(0, 80) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = "none"
  )

```
#### 4.1.2 FIGURE 5a: Subgroup Analysis - for each subgroup

```{r}
# Build dummy matrix once (no intercept)
X_full <- model.matrix(~ age_group + gender + season, data = df)[, -1]

# Helper: builds X for a subset, drops any constant columns
build_X_subset <- function(subset_df) {
  X_sub <- model.matrix(~ age_group + gender + season, data = subset_df)[, -1]
  # Drop columns that are constant within the subset (not informative)
  X_sub <- X_sub[, apply(X_sub, 2, var) != 0, drop = FALSE]
  return(X_sub)
}

# Function to run causal forest on a subset
run_causal_forest <- function(subset_df) {
  X_sub <- build_X_subset(subset_df)
  
  cf_model <- causal_forest(
    X = X_sub,
    Y = subset_df$value_result,
    W = subset_df$treatment,
    num.trees = 2000
  )
  
  # ATE estimate and CI
  ate <- average_treatment_effect(cf_model)
  ate_est <- ate[1]
  ate_se <- ate[2]
  
  z_99 <- 2.576
  z_95 <- 1.960
  
  ci_99 <- c(ate_est - z_99 * ate_se, ate_est + z_99 * ate_se)
  ci_95 <- c(ate_est - z_95 * ate_se, ate_est + z_95 * ate_se)
  
  avg_pre <- 26.3  # or calculate dynamically
  pct_change <- (ate_est / avg_pre) * 100
  
  data.frame(
    Category = "",
    ATE_Estimate = round(ate_est, 4),
    CI_Lower_99 = round(ci_99[1], 4),
    CI_Upper_99 = round(ci_99[2], 4),
    CI_Lower_95 = round(ci_95[1], 4),
    CI_Upper_95 = round(ci_95[2], 4),
    Percentage_Change = round(pct_change, 2)
  )
}

# Run subgroup analyses
age_group_results <- lapply(unique(df$age_group), function(age) {
  sub <- df[df$age_group == age, ]
  res <- run_causal_forest(sub)
  res$Category <- paste("Age Group:", age)
  res
})

gender_results <- lapply(unique(df$gender), function(g) {
  sub <- df[df$gender == g, ]
  res <- run_causal_forest(sub)
  res$Category <- paste("Gender:", g)
  res
})

season_results <- lapply(unique(df$season), function(s) {
  sub <- df[df$season == s, ]
  res <- run_causal_forest(sub)
  res$Category <- paste("Season:", s)
  res
})

# Combine all results
final_results <- do.call(rbind, c(age_group_results, gender_results, season_results))
print(final_results)

```

```{r}
# Adjust the order for seasons
season_order <- c("Season: Fall", "Season: Summer", "Season: Spring", "Season: Winter")

# Set factor levels for the Category column
final_results$Category <- factor(
  final_results$Category,
  levels = c(season_order, setdiff(unique(final_results$Category), season_order))
)

forest_plot <- ggplot(final_results, aes(x = Category, y = ATE_Estimate)) +
  # Add the points for ATE estimates
  geom_point(size = 5, shape = 21, fill = "black", color = "black", stroke = 0.5) +
  
  # Add 99% CI as the longest bars
  geom_linerange(aes(ymin = CI_Lower_99, ymax = CI_Upper_99), size = 1.0, color = "black") +
  
  # Add 95% CI as shorter bars
  geom_linerange(aes(ymin = CI_Lower_95, ymax = CI_Upper_95), size = 2.0, color = "firebrick") +
  
  # Add a dashed line at 0 for reference
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  
  # Flip coordinates for horizontal plot
  coord_flip() +
  
  # Labels and theme
  labs(
    x = "Subgroup category",
    y = "Estimated reduction (in µg/l)"  
  ) +
  theme_classic(base_size = 16) +  # Increased base font size
  theme(
    axis.text.x = element_text(hjust = 0.5, size = 14),  
    axis.text.y = element_text(size = 14),
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_blank()
  )

# Display the updated plot
print(forest_plot)
```

#### 4.1.3 FIGURE 4a: Comparison of effect sizes across all 3 methods used
```{r}
# Cohens d and SMD are standardized, transform back by multiplying with pooled SD

# Calculate pooled SD for original data
s_pooled <- sqrt(((n1 - 1) * sd_b^2 + (n2 - 1) * sd_d^2) / (n1 + n2 - 2))

# ATE from Cohen's d
ate_from_cohens_d <- cohens_d * s_pooled

# ATE confidence intervals for Cohen's d (95% instead of 90%)
ci_99_lower_ate_d <- ci_99_lower_d * s_pooled
ci_99_upper_ate_d <- ci_99_upper_d * s_pooled
ci_95_lower_ate_d <- ci_95_lower_d * s_pooled
ci_95_upper_ate_d <- ci_95_upper_d * s_pooled

# ATE from SMD
ate_from_smd <- smd * s_pooled

# ATE confidence intervals for SMD (95% instead of 90%)
ci_99_lower_ate_smd <- ci_99_lower_smd * s_pooled
ci_99_upper_ate_smd <- ci_99_upper_smd * s_pooled
ci_95_lower_ate_smd <- ci_95_lower_smd * s_pooled
ci_95_upper_ate_smd <- ci_95_upper_smd * s_pooled

# Create updated data frame with recalculated ATE values
effect_sizes_updated <- data.frame(
  Category = c("Cohen's d", "SMD (PSM)", "ATE (ML)"),
  Estimate = c(ate_from_cohens_d, ate_from_smd, ate_estimate),
  CI_Lower_99 = c(ci_99_lower_ate_d, ci_99_lower_ate_smd, ci_99_lower_ate),
  CI_Upper_99 = c(ci_99_upper_ate_d, ci_99_upper_ate_smd, ci_99_upper_ate),
  CI_Lower_95 = c(ci_95_lower_ate_d, ci_95_lower_ate_smd, ci_95_lower_ate),
  CI_Upper_95 = c(ci_95_upper_ate_d, ci_95_upper_ate_smd, ci_95_upper_ate)
)

# Reorder the Category levels to match the new desired order (Cohen’s d on top, ATE ML at bottom)
effect_sizes_updated$Category <- factor(effect_sizes_updated$Category, levels = c("Cohen's d", "SMD (PSM)", "ATE (ML)"))

# Create the forest plot with overlapping whiskers for CIs
forest_plot_updated <- ggplot(effect_sizes_updated, aes(x = Category, y = Estimate)) +
  # Add points for the ATE estimates
  geom_point(size = 5, shape = 21, fill = "black", color = "black", stroke = 0.5) +
  
  # Add 99% CI as the longest bars
  geom_linerange(aes(ymin = CI_Lower_99, ymax = CI_Upper_99), size = 1.0, color = "black") +
  
  # Add 95% CI as shorter bars
  geom_linerange(aes(ymin = CI_Lower_95, ymax = CI_Upper_95), size = 2.0, color = "firebrick") +
  
  # Add a dashed line at 0 for reference
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  
  # Flip coordinates for horizontal orientation
  coord_flip() +
  
  # Labels and theme
  labs(
    y = "Estimated effect size (in µg/l)",  # Updated y-axis label
    x = NULL  # Remove x-axis label
  ) +
  theme_classic(base_size = 16) +  # Set a clean, classic theme
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    plot.title = element_text(size = 18, face = "bold"),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_blank()
  )

# Display the updated plot
print(forest_plot_updated)

```



#### 4.1.4 Sensitivity analysis with sensemakr

```{r}
# since sensemakr works directly with linear regression models, we need to approximate the causal forest results in a linear regression framework

# Dummy encode with cleaned column names
X_lm <- model.matrix(~ age + gender + season, data = df)[, -1]

# Clean up problematic characters in column names
colnames(X_lm) <- make.names(colnames(X_lm))

# Combine into final data frame
df_linear <- cbind(value_result = df$value_result, treatment = df$treatment, X_lm)

# Build the formula dynamically
predictor_vars <- colnames(df_linear)[-1]  # exclude response
formula_str <- paste("value_result ~", paste(predictor_vars, collapse = " + "))
formula_obj <- as.formula(formula_str)

# Fit the linear regression model
linear_model <- lm(formula_obj, data = as.data.frame(df_linear))

# Summarize
summary(linear_model)


```

```{r}
# Identify which covariates to use as benchmarks (all except treatment)
benchmark_vars <- setdiff(colnames(df_linear), c("value_result", "treatment"))

# Run sensitivity analysis with the correct variable names
sensitivity_analysis <- sensemakr(
  model = linear_model,
  treatment = "treatment", 
  benchmark_covariates = benchmark_vars,  # use actual dummy names
  kd = 1
)

# Summarize results
summary(sensitivity_analysis)
```
#### 4.1.5 Weighted ASMD Analysis
```{r}
df$period <- as.factor(df$period)

# Convert categorical variables to numeric for mean calculation
df <- df %>%
  mutate(
    age_group_numeric = as.numeric(as.factor(age_group)),  # Encodes age groups as 1, 2, 3
    gender_numeric = as.numeric(as.factor(gender)),        # Encodes gender (F=1, M=2)
    test_month_numeric = as.numeric(as.factor(test_month))  # Encodes months as 1-12
  )

# Estimate Propensity Scores
ps_model <- glm(period ~ age_group + gender + test_month, data = df, family = binomial)
df$pscore <- predict(ps_model, type = "response")  

# Compute Inverse Probability of Treatment Weights (IPTW)
df$iptw <- ifelse(df$period == "D", 1 / df$pscore, 1 / (1 - df$pscore))

# Compute Balance Statistics Using Weighted Data
balance_stats <- bal.tab(ps_model, weights = df$iptw, 
                         covs = df[, c("age_group", "gender", "test_month")], 
                         estimand = "ATE", thresholds = c(m = 0.1)) 

# Convert to Data Frame
asmd_results <- as.data.frame(balance_stats$Balance)

# Compute Mean ASMD for Each Category
asmd_summary <- data.frame(
  Covariate = c("Age Group", "Gender", "Test Month"),
  Weighted_ASMD = c(  # Fix column name issue
    mean(asmd_results[grep("age_group", rownames(asmd_results)), "Diff.Adj"], na.rm = TRUE),
    mean(asmd_results[grep("gender", rownames(asmd_results)), "Diff.Adj"], na.rm = TRUE),
    mean(asmd_results[grep("test_month", rownames(asmd_results)), "Diff.Adj"], na.rm = TRUE)
  ),
  M.Threshold = "Balanced, <0.1"
)

# Fix Covariate Naming to Match `asmd_results`
means_table <- df %>%
  group_by(period) %>%
  summarise(
    `Age Group` = mean(age_group_numeric, na.rm = TRUE),
    `Gender` = mean(gender_numeric, na.rm = TRUE),
    `Test Month` = mean(test_month_numeric, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -period, names_to = "Covariate", values_to = "Mean")

# Reshape Table to Wide Format (Pre-Pandemic & Pandemic Means)
means_table_wide <- means_table %>%
  pivot_wider(names_from = period, values_from = Mean) %>%
  rename(`Pre-Pandemic Mean` = `B`, `Pandemic Mean` = `D`)

# Merge Means with ASMD Data
final_balance_table <- asmd_summary %>%
  rename(`Weighted ASMD` = Weighted_ASMD) %>%  # Ensure correct column name
  left_join(means_table_wide, by = "Covariate") %>%
  dplyr::select(Covariate, `Pre-Pandemic Mean`, `Pandemic Mean`, `Weighted ASMD`, M.Threshold)


# Print Final Table for Review
print(final_balance_table)
  
# Categorical breakdown
# Extract and format the full ASMD table (with all individual levels)
asmd_detailed <- asmd_results %>%
  rownames_to_column("Covariate_Level") %>%
  dplyr::select(Covariate_Level, Diff.Adj) %>%
  rename(`Weighted ASMD` = Diff.Adj)

# Print head for review
print(head(asmd_detailed, 10))

# Compute variance ratio of propensity scores between treatment groups
var_ps_treated <- var(df$pscore[df$period == "D"], na.rm = TRUE)
var_ps_control <- var(df$pscore[df$period == "B"], na.rm = TRUE)
variance_ratio <- var_ps_treated / var_ps_control

cat("Variance Ratio (Treated / Control):", round(variance_ratio, 3), "\n")
  
```

### 4.2 Change in vitamin D deficiency rates

```{r}
# 1. Treatment indicator (already binary)
df$treatment <- ifelse(df$period == "D", 1, 0)

# 2. Ensure the deficiency column is numeric 
df$deficient <- as.numeric(df$deficient)

# 3. Create design matrix (one-hot encode categorical variables)
X_def <- model.matrix(~ age_group + gender + season, data = df)[, -1]  # remove intercept

# 4. Calculate pre-pandemic average deficiency rate
pre_pandemic_deficiency_rate <- df %>%
  filter(period == "B") %>%
  summarise(avg_deficiency = mean(deficient, na.rm = TRUE)) %>%
  pull(avg_deficiency)

cat("Pre-pandemic average deficiency rate:", round(pre_pandemic_deficiency_rate, 4), "\n")

# 5. Run causal forest model
cf_deficiency <- causal_forest(
  X = X_def,
  Y = df$deficient,
  W = df$treatment,
  num.trees = 2000
)

# 6. Estimate ATE
ate_result_deficiency <- average_treatment_effect(cf_deficiency, target.sample = "all")
ate_estimate_deficiency <- ate_result_deficiency[1]
ate_std_err_deficiency <- ate_result_deficiency[2]

# 7. Print ATE
cat("ATE Estimate (Effect on Deficiency Rate):", round(ate_estimate_deficiency, 4), "\n")

# 8. Percentage change relative to pre-pandemic rate
percentage_change_deficiency <- (ate_estimate_deficiency / pre_pandemic_deficiency_rate) * 100
cat("Percentage change due to pandemic:", round(percentage_change_deficiency, 2), "%\n")

# 9. Confidence intervals
z_99 <- 2.576
z_95 <- 1.960 

ci_99_lower_ate_def <- ate_estimate_deficiency - z_99 * ate_std_err_deficiency
ci_99_upper_ate_def <- ate_estimate_deficiency + z_99 * ate_std_err_deficiency

ci_95_lower_ate_def <- ate_estimate_deficiency - z_95 * ate_std_err_deficiency
ci_95_upper_ate_def <- ate_estimate_deficiency + z_95 * ate_std_err_deficiency

cat("99% CI for ATE (Deficiency Rate):", round(ci_99_lower_ate_def, 4), "to", round(ci_99_upper_ate_def, 4), "\n")
cat("95% CI for ATE (Deficiency Rate):", round(ci_95_lower_ate_def, 4), "to", round(ci_95_upper_ate_def, 4), "\n")

```

#### 4.2.1 FIGURE 4b: Comparison of effect sizes across all 3 methods used

```{r}
# Extracting ATE estimate and standard error from causal forest model
ate_estimate_def <- ate_result_deficiency[1]
ate_std_err_def <- ate_result_deficiency[2]

# Confidence intervals for ATE
ci_99_lower_ate_def <- ate_estimate_def - z_99 * ate_std_err_def
ci_99_upper_ate_def <- ate_estimate_def + z_99 * ate_std_err_def
ci_95_lower_ate_def <- ate_estimate_def - z_95 * ate_std_err_def
ci_95_upper_ate_def <- ate_estimate_def + z_95 * ate_std_err_def

# Bring all effect size metrics together in one df
effect_sizes_deficiency <- data.frame(
  Category = c("Cohen's d", "SMD (PSM)", "ATE (ML)"),
  Estimate = c(cohens_d_def * sd_pooled_def, smd_def * sd_matched_pooled_def, ate_estimate_def),
  CI_Lower_99 = c(ci_99_lower_d_def * sd_pooled_def, ci_99_lower_smd_def * sd_matched_pooled_def, ci_99_lower_ate_def),
  CI_Upper_99 = c(ci_99_upper_d_def * sd_pooled_def, ci_99_upper_smd_def * sd_matched_pooled_def, ci_99_upper_ate_def),
  CI_Lower_95 = c(ci_95_lower_d_def * sd_pooled_def, ci_95_lower_smd_def * sd_matched_pooled_def, ci_95_lower_ate_def),
  CI_Upper_95 = c(ci_95_upper_d_def * sd_pooled_def, ci_95_upper_smd_def * sd_matched_pooled_def, ci_95_upper_ate_def)
)

# Reorder factor levels for display order
effect_sizes_deficiency$Category <- factor(effect_sizes_deficiency$Category, levels = c("Cohen's d", "SMD (PSM)", "ATE (ML)"))

forest_plot_deficiency <- ggplot(effect_sizes_deficiency, aes(x = Category, y = Estimate)) +
  # Add points for the ATE estimates
  geom_point(size = 5, shape = 21, fill = "black", color = "black", stroke = 0.5) +
  
  # Add 99% CI as the longest bars
  geom_linerange(aes(ymin = CI_Lower_99, ymax = CI_Upper_99), size = 1.0, color = "black") +
  
  # Add 95% CI as shorter bars
  geom_linerange(aes(ymin = CI_Lower_95, ymax = CI_Upper_95), size = 2.0, color = "firebrick") +
  
  # Add a dashed line at 0 for reference
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  
  # Flip coordinates for horizontal orientation
  coord_flip() +
  
  # Labels and theme
  labs(
    y = "Estimated effect size (percentage points)",  
    x = NULL  
  ) +
  theme_classic(base_size = 16) +  
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    plot.title = element_text(size = 18, face = "bold"),
    panel.grid.major.y = element_line(color = "grey80"),
    panel.grid.minor.y = element_blank()
  )

# Display the updated plot
print(forest_plot_deficiency)

```


## 5. Negative Controls Outcome Analysis 1: Entire analysis on a different time frame

```{r}
# Define the new periods - uncomment the desired time frame to be tested

# Time Frame 1
period1 <- c(as.Date("2018-03-01"), as.Date("2020-01-31"))
period2 <- c(as.Date("2020-02-01"), as.Date("2022-02-28"))

# Time Frame 2
#period1 <- c(as.Date("2018-03-01"), as.Date("2020-03-31"))
#period2 <- c(as.Date("2020-04-01"), as.Date("2022-02-28"))

# Create a new data frame for robustness check
df_robust <- df

# Assign each entry to a period
assign_period <- function(date) {
  if (date >= period1[1] && date <= period1[2]) {
    return("B")
  } else if (date >= period2[1] && date <= period2[2]) {
    return("D")
  } else {
    return(NA) 
  }
}

df_robust$period <- as.character(sapply(df_robust$test_date, assign_period))
df_robust <- df_robust[!is.na(df_robust$period),]
```


### 5.1 Descriptive
```{r}
sample_size_period <- df_robust %>% count(period)
print("Sample size per period:")
print(sample_size_period)

sample_size_contingency <- with(df_robust, table(period))
sample_size_p_value <- chisq.test(sample_size_contingency)$p.value
print("P-value for sample size distribution:")
print(sample_size_p_value)

age_group_counts <- df_robust %>% group_by(period, age_group) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print("Age group counts per period:")
print(age_group_counts)

age_group_p_value <- chisq.test(with(df_robust, table(period, age_group)))$p.value
print("P-value for age group distribution:")
print(age_group_p_value)

gender_group_counts <- df_robust %>% group_by(period, gender) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print("Gender group counts per period:")
print(gender_group_counts)

gender_group_p_value <- chisq.test(with(df_robust, table(period, gender)))$p.value
print("P-value for gender distribution:")
print(gender_group_p_value)

season_counts <- df_robust %>% group_by(period, season) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print("Seasonal distribution per period:")
print(season_counts)

season_counts_p_value <- chisq.test(with(df_robust, table(period, season)))$p.value
print("P-value for seasonal distribution:")
print(season_counts_p_value)

mean_vitamin_d_levels <- df_robust %>% group_by(period) %>%
  summarise(mean_value = mean(value_result, na.rm = TRUE),
            sd_value = sd(value_result, na.rm = TRUE),
            .groups = 'drop')
print("Mean and standard deviation of Vitamin D levels per period:")
print(mean_vitamin_d_levels)

vitamin_d_t_test <- t.test(value_result ~ period, data = df_robust)
print("T-test results for Vitamin D levels:")
print(vitamin_d_t_test)

vitamin_d_deficiency_rate <- df_robust %>% group_by(period) %>%
  summarise(deficiency_count = sum(deficient), 
            total_count = n(), 
            deficiency_rate = sum(deficient) / n() * 100, 
            .groups = 'drop')
print("Vitamin D deficiency rates per period:")
print(vitamin_d_deficiency_rate)

deficiency_contingency <- with(df_robust, table(period, deficient))
deficiency_p_value <- chisq.test(deficiency_contingency)$p.value
print("P-value for Vitamin D deficiency rate differences:")
print(deficiency_p_value)

```

### 5.2 PSM

```{r}
df_robust$gender <- factor(df_robust$gender)
df_robust$age_group <- factor(df_robust$age_group)
df_robust$period <- factor(df_robust$period)
df_robust$test_month <- factor(df_robust$test_month)

test_match_obj <- matchit(period ~ age_group + gender + test_month, data = df_robust, 
                          method = "nearest", distance = "glm", 
                          exact = c("gender", "age_group", "test_month")) 
matched_data <- match.data(test_match_obj)

# Statistical testing for matched data
vitamin_d_t_test_matched <- t.test(value_result ~ period, data = matched_data)
print("T-test results for Vitamin D levels in matched data:")
print(vitamin_d_t_test_matched)

# Contingency table for deficiency rates
deficiency_contingency_matched <- with(matched_data, table(period, deficient))
deficiency_p_value_matched <- chisq.test(deficiency_contingency_matched)$p.value
print("P-value for Vitamin D deficiency rate differences in matched data:")
print(deficiency_p_value_matched)

# Calculate deficiency rates before and during the pandemic
deficiency_rates_matched <- prop.table(deficiency_contingency_matched, margin = 1) * 100
print("Deficiency rates in matched sample (before and during the pandemic):")
print(deficiency_rates_matched)

```

### 5.3 ML

```{r}
# Create treatment indicator
df_robust$treatment <- ifelse(df_robust$period == "D", 1, 0)

# Use model.matrix to one-hot encode categorical covariates (drop intercept column)
X <- model.matrix(~ age_group + gender + season, data = df_robust)[, -1]

# Fit causal forest
causal_forest_model <- causal_forest(
  X = X, 
  Y = df_robust$value_result, 
  W = df_robust$treatment, 
  num.trees = 2000
)

# Estimate ATE
ate_result <- average_treatment_effect(causal_forest_model, target.sample = "all")

# Extract point estimate and standard error
ate_estimate <- ate_result[1]
ate_std_err <- ate_result[2]

# Print ATE
print(paste("ATE Estimate:", round(ate_estimate, 4)))

# Calculate and print percentage change
average_value_result <- mean(df_robust$value_result, na.rm = TRUE)
percentage_change <- (ate_estimate / average_value_result) * 100
print(paste("Percentage change due to pandemic:", round(percentage_change, 2), "%"))

# Confidence intervals
z_99 <- qnorm(0.995)
z_95 <- qnorm(0.975)

ci_99_lower_ate <- ate_estimate - z_99 * ate_std_err
ci_99_upper_ate <- ate_estimate + z_99 * ate_std_err
ci_95_lower_ate <- ate_estimate - z_95 * ate_std_err
ci_95_upper_ate <- ate_estimate + z_95 * ate_std_err

# Output confidence intervals
cat("99% Confidence Interval for ATE:", ci_99_lower_ate, "to", ci_99_upper_ate, "\n")
cat("95% Confidence Interval for ATE:", ci_95_lower_ate, "to", ci_95_upper_ate, "\n")

# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.
```

## 6. Robustness check: Compare different months of different years
### 6.1 Descriptive
```{r}
# Filter data for the months of March, April, and May for the years 2018, 2019, 2020, and 2021
df_spring <- df %>%
  filter(year(test_date) %in% c(2018, 2019, 2020, 2021) &
           month(test_date) %in% c(3, 4, 5))

# Ensure factors
df_spring$period <- factor(df_spring$period)
df_spring$gender <- factor(df_spring$gender)
df_spring$age_group <- factor(df_spring$age_group)
df_spring$season <- factor(df_spring$season)
df_spring$test_month <- factor(month(df_spring$test_date))

# 1. Descriptive Statistics
spring_vitamin_d_stats <- df_spring %>%
  group_by(cohort_year, test_month) %>%
  summarise(mean_vit_d = mean(value_result, na.rm = TRUE),
            sd_vit_d = sd(value_result, na.rm = TRUE),
            deficiency_count = sum(deficient, na.rm = TRUE),
            total_count = n(),
            deficiency_rate = (sum(deficient, na.rm = TRUE) / n()) * 100,
            .groups = "drop") %>%
  arrange(test_month, cohort_year)

print("Vitamin D statistics by year and month:")
print(spring_vitamin_d_stats)

# Mean Vitamin D levels before and during the pandemic
mean_vit_d_period <- df_spring %>%
  group_by(period) %>%
  summarise(mean_vit_d = mean(value_result, na.rm = TRUE),
            sd_vit_d = sd(value_result, na.rm = TRUE),
            .groups = "drop")

print("Mean Vitamin D levels before and during the pandemic:")
print(mean_vit_d_period)

# Deficiency rates before and during the pandemic
deficiency_rate_period <- df_spring %>%
  group_by(period) %>%
  summarise(deficiency_count = sum(deficient, na.rm = TRUE),
            total_count = n(),
            deficiency_rate = (sum(deficient, na.rm = TRUE) / n()) * 100,
            .groups = "drop")

print("Deficiency rates before and during the pandemic:")
print(deficiency_rate_period)

# ANOVA test for mean Vitamin D levels
vitamin_d_anova <- aov(value_result ~ factor(cohort_year) * factor(test_month), data = df_spring)
summary(vitamin_d_anova)

# Chi-square test for deficiency rate
deficiency_contingency <- with(df_spring, table(cohort_year, test_month, deficient))
deficiency_p_value <- chisq.test(rowSums(deficiency_contingency))$p.value
print("P-value for Vitamin D deficiency rate differences across years and months:")
print(deficiency_p_value)

```
### 6.2 PSM
```{r}
test_match_obj <- matchit(period ~ age_group + gender + test_month, data = df_spring, 
                          method = "nearest", distance = "glm", 
                          exact = c("gender", "age_group", "test_month"))
matched_data <- match.data(test_match_obj)

# T-test for Vitamin D levels in matched data
vitamin_d_t_test_matched <- t.test(value_result ~ period, data = matched_data)
print("T-test results for Vitamin D levels in matched data:")
print(vitamin_d_t_test_matched)

# Deficiency rates in matched data
deficiency_contingency_matched <- with(matched_data, table(period, deficient))
deficiency_p_value_matched <- chisq.test(deficiency_contingency_matched)$p.value
print("P-value for Vitamin D deficiency rate differences in matched data:")
print(deficiency_p_value_matched)

deficiency_rates_matched <- prop.table(deficiency_contingency_matched, margin = 1) * 100
print("Deficiency rates in matched sample (before and during the pandemic):")
print(deficiency_rates_matched)
```

### 6.3 ML
```{r}

# Create treatment indicator
df_spring$treatment <- ifelse(df_spring$period == "D", 1, 0)

# Ensure categorical variables are factors
df_spring$age_group <- as.factor(df_spring$age_group)
df_spring$gender <- as.factor(df_spring$gender)
df_spring$season <- as.factor(df_spring$season)

# Drop any factor with only one level
covariates <- c("age_group", "gender", "season")
valid_covariates <- covariates[sapply(df_spring[covariates], function(x) length(unique(x)) > 1)]

# Build formula dynamically based on valid covariates
formula <- as.formula(paste("~", paste(valid_covariates, collapse = " + ")))

# Generate model matrix
X_spring <- model.matrix(formula, data = df_spring)[, -1]


# Fit causal forest
causal_forest_model <- causal_forest(
  X = X_spring,
  Y = df_spring$value_result,
  W = df_spring$treatment,
  num.trees = 2000
)

# Estimate ATE
ate_result <- average_treatment_effect(causal_forest_model, target.sample = "all")
ate_estimate <- ate_result[1]
ate_std_err <- ate_result[2]

# Print results
print(paste("ATE Estimate:", round(ate_estimate, 4)))

# Percentage change based on pre-pandemic average
average_value_result <- mean(df_spring$value_result, na.rm = TRUE)
percentage_change <- (ate_estimate / average_value_result) * 100
print(paste("Percentage change due to pandemic:", round(percentage_change, 2), "%"))

# Confidence intervals
z_99 <- qnorm(0.995)
z_95 <- qnorm(0.975)

ci_99_lower_ate <- ate_estimate - z_99 * ate_std_err
ci_99_upper_ate <- ate_estimate + z_99 * ate_std_err
ci_95_lower_ate <- ate_estimate - z_95 * ate_std_err
ci_95_upper_ate <- ate_estimate + z_95 * ate_std_err

# Output CI results
cat("99% Confidence Interval for ATE:", ci_99_lower_ate, "to", ci_99_upper_ate, "\n")
cat("95% Confidence Interval for ATE:", ci_95_lower_ate, "to", ci_95_upper_ate, "\n")


# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.

```


## 7. Robustness check: Hours of sunshine

```{r}
# Define the base URL
base_url <- "https://opendata.dwd.de/climate_environment/CDC/regional_averages_DE/monthly/sunshine_duration/"

# Initialize an empty data frame to store sunshine data
sunshine_data <- data.frame()

# Loop through each month to read and combine the data
for (i in 1:12) {
  # Format the file name
  file_name <- sprintf("regional_averages_sd_%02d.txt", i)
  file_url <- paste0(base_url, file_name)
  
  # Read the data, skipping the first row to use the second row as headers
  monthly_data <- read_delim(file_url, delim = ";", col_types = cols(), skip = 1)
  
  # Select relevant columns (Jahr = year, Monat = month, Bayern = sunshine duration)
  monthly_data <- monthly_data %>%
    dplyr::select(Jahr, Monat, Bayern) %>%
    rename(sunshine_duration = Bayern) %>%
    mutate(Monat = as.numeric(Monat)) # Ensure month is numeric
  
  # Append the monthly data to the combined data frame
  sunshine_data <- bind_rows(sunshine_data, monthly_data)
}

# Ensuring year and month are numeric for merging
df <- df %>%
  mutate(
    cohort_year = as.numeric(cohort_year),
    test_month = as.numeric(test_month)
  )

# Merge the sunshine data with your existing data frame
df <- df %>%
  left_join(sunshine_data, by = c("cohort_year" = "Jahr", "test_month" = "Monat"))

# Check for any unmatched entries
unmatched_entries <- df %>%
  filter(is.na(sunshine_duration))

# Print the number of unmatched entries, if any
cat("Number of entries with missing sunshine duration data:", nrow(unmatched_entries), "\n")

# View the first few rows of the updated data frame
head(df)

```

```{r}
# Create treatment indicator
df$treatment <- ifelse(df$period == "D", 1, 0)

# Ensure sunshine_duration is numeric (already handled)
df$sunshine_numeric <- as.numeric(df$sunshine_duration)

# One-hot encode categorical variables with model.matrix and exclude intercept
X_categorical <- model.matrix(~ age_group + gender + season, data = df)[, -1]

# Combine with sunshine variable
X <- cbind(X_categorical, sunshine_numeric = df$sunshine_numeric)

# Fit causal forest with sunshine covariate
causal_forest_model_sunshine <- causal_forest(
  X = X, 
  Y = df$value_result, 
  W = df$treatment, 
  num.trees = 2000
)

# Estimate ATE (target.sample = "overlap" due to extreme propensity scores with sunshine)
ate_result_sunshine <- average_treatment_effect(causal_forest_model_sunshine, target.sample = "overlap")
ate_estimate_sunshine <- ate_result_sunshine[1]
ate_std_err_sunshine <- ate_result_sunshine[2]

# Print ATE estimate
print(paste("ATE Estimate (with sunshine):", round(ate_estimate_sunshine, 4)))

# Confidence intervals
z_99 <- 2.576
z_95 <- 1.960

ci_99_lower_ate_sunshine <- ate_estimate_sunshine - z_99 * ate_std_err_sunshine
ci_99_upper_ate_sunshine <- ate_estimate_sunshine + z_99 * ate_std_err_sunshine

ci_95_lower_ate_sunshine <- ate_estimate_sunshine - z_95 * ate_std_err_sunshine
ci_95_upper_ate_sunshine <- ate_estimate_sunshine + z_95 * ate_std_err_sunshine

# Print confidence intervals
cat("99% CI for ATE (with sunshine):", ci_99_lower_ate_sunshine, "to", ci_99_upper_ate_sunshine, "\n")
cat("95% CI for ATE (with sunshine):", ci_95_lower_ate_sunshine, "to", ci_95_upper_ate_sunshine, "\n")

# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.

```

## 8. Robustness check: Randomly shuffling the Vit D results

```{r}
# Create a copy of df with shuffled Vitamin D values
df_shuffled <- df %>%
  mutate(shuffled_value = sample(value_result, replace = FALSE))

# Ensure there are no missing values in the shuffled data
df_shuffled <- df_shuffled %>%
  filter(!is.na(shuffled_value))
```

### 8.1 Descriptive
```{r}
# Compute statistics for the shuffled Vitamin D values
vitamin_d_stats_shuffled <- df_shuffled %>%
  group_by(cohort_year, test_month) %>%
  summarise(mean_vit_d = mean(shuffled_value, na.rm = TRUE),
            sd_vit_d = sd(shuffled_value, na.rm = TRUE),
            deficiency_count = sum(shuffled_value < thresholds$VD3, na.rm = TRUE),
            total_count = n(),
            deficiency_rate = (sum(shuffled_value < thresholds$VD3, na.rm = TRUE) / n()) * 100,
            .groups = "drop") %>%
  arrange(test_month, cohort_year)

print("Shuffled Vitamin D statistics by year and month:")
print(vitamin_d_stats_shuffled)

# Mean Vitamin D levels before and during the pandemic
mean_vit_d_period_shuffled <- df_shuffled %>%
  group_by(period) %>%
  summarise(mean_vit_d = mean(shuffled_value, na.rm = TRUE),
            sd_vit_d = sd(shuffled_value, na.rm = TRUE),
            .groups = "drop")

print("Mean Shuffled Vitamin D levels before and during the pandemic:")
print(mean_vit_d_period_shuffled)

# Deficiency rates before and during the pandemic
deficiency_rate_period_shuffled <- df_shuffled %>%
  group_by(period) %>%
  summarise(deficiency_count = sum(shuffled_value < thresholds$VD3, na.rm = TRUE),
            total_count = n(),
            deficiency_rate = (sum(shuffled_value < thresholds$VD3, na.rm = TRUE) / n()) * 100,
            .groups = "drop")

print("Shuffled Deficiency rates before and during the pandemic:")
print(deficiency_rate_period_shuffled)

# ANOVA test for mean Vitamin D levels on shuffled data
vitamin_d_anova_shuffled <- aov(shuffled_value ~ factor(cohort_year) * factor(test_month), data = df_shuffled)
summary(vitamin_d_anova_shuffled)

# Chi-square test for deficiency rate in shuffled data
deficiency_contingency_shuffled <- with(df_shuffled, table(cohort_year, test_month, shuffled_value < thresholds$VD3))
deficiency_p_value_shuffled <- chisq.test(rowSums(deficiency_contingency_shuffled))$p.value
print("P-value for Shuffled Vitamin D deficiency rate differences across years and months:")
print(deficiency_p_value_shuffled)
```

### 8.2 PSM
```{r}
test_match_obj_shuffled <- matchit(period ~ age_group + gender + test_month, data = df_shuffled, 
                                   method = "nearest", distance = "glm", 
                                   exact = c("gender", "age_group", "test_month"))
matched_data_shuffled <- match.data(test_match_obj_shuffled)

# T-test for Vitamin D levels in matched shuffled data
vitamin_d_t_test_matched_shuffled <- t.test(shuffled_value ~ period, data = matched_data_shuffled)
print("T-test results for Shuffled Vitamin D levels in matched data:")
print(vitamin_d_t_test_matched_shuffled)

# Deficiency rates in matched shuffled data
deficiency_contingency_matched_shuffled <- with(matched_data_shuffled, table(period, shuffled_value < thresholds$VD3))
deficiency_p_value_matched_shuffled <- chisq.test(deficiency_contingency_matched_shuffled)$p.value
print("P-value for Shuffled Vitamin D deficiency rate differences in matched data:")
print(deficiency_p_value_matched_shuffled)

deficiency_rates_matched_shuffled <- prop.table(deficiency_contingency_matched_shuffled, margin = 1) * 100
print("Shuffled Deficiency rates in matched sample (before and during the pandemic):")
print(deficiency_rates_matched_shuffled)
```


### 8.3 ML
```{r}
# Create treatment indicator
df_shuffled$treatment <- ifelse(df_shuffled$period == "D", 1, 0)

# One-hot encode categorical variables using model.matrix, removing intercept
X_shuffled <- model.matrix(~ age_group + gender + season, data = df_shuffled)[, -1]

# Run causal forest with shuffled outcome
causal_forest_model_shuffled <- causal_forest(
  X = X_shuffled, 
  Y = df_shuffled$shuffled_value, 
  W = df_shuffled$treatment, 
  num.trees = 2000
)

# Estimate ATE
ate_result_shuffled <- average_treatment_effect(causal_forest_model_shuffled, target.sample = "all")
ate_estimate_shuffled <- ate_result_shuffled[1]
ate_std_err_shuffled <- ate_result_shuffled[2]

# Output ATE estimate
print(paste("Shuffled ATE Estimate:", round(ate_estimate_shuffled, 4)))

# Compute percentage change
average_value_result_shuffled <- mean(df_shuffled$shuffled_value, na.rm = TRUE)
percentage_change_shuffled <- (ate_estimate_shuffled / average_value_result_shuffled) * 100
print(paste("Shuffled Percentage change due to pandemic:", round(percentage_change_shuffled, 2), "%"))

# Confidence intervals
z_99 <- 2.576
z_95 <- 1.960

ci_99_lower_ate_shuffled <- ate_estimate_shuffled - z_99 * ate_std_err_shuffled
ci_99_upper_ate_shuffled <- ate_estimate_shuffled + z_99 * ate_std_err_shuffled
ci_95_lower_ate_shuffled <- ate_estimate_shuffled - z_95 * ate_std_err_shuffled
ci_95_upper_ate_shuffled <- ate_estimate_shuffled + z_95 * ate_std_err_shuffled

cat("99% Confidence Interval for Shuffled ATE:", ci_99_lower_ate_shuffled, "to", ci_99_upper_ate_shuffled, "\n")
cat("95% Confidence Interval for Shuffled ATE:", ci_95_lower_ate_shuffled, "to", ci_95_upper_ate_shuffled, "\n")


# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.
```

-------------------

## 9. New analyses that were added in response to the reviewer's comments

### 9.1 Addition of postal code as covariate in PSM and ML

#### 9.1.0 Exploration & Descriptive Analyses with the postal code
```{r}
# Descriptive statistics with PLZ mapping

# 1. Unique patients per postal code
patients_per_postal_code <- df %>%
  group_by(postal_code) %>%
  summarise(unique_patients = n_distinct(patient), .groups = 'drop')

# 2. Unique patients per PLZ mapping value
patients_per_plz_mapping <- df %>%
  group_by(plz_mapping) %>%
  summarise(unique_patients = n_distinct(patient), .groups = 'drop')

# 3. Unique patients per age group and PLZ mapping
patients_by_age_plz <- df %>%
  group_by(age_group, plz_mapping) %>%
  summarise(unique_patients = n_distinct(patient), .groups = 'drop')

# 4. Unique patients per gender and PLZ mapping
patients_by_gender_plz <- df %>%
  group_by(gender, plz_mapping) %>%
  summarise(unique_patients = n_distinct(patient), .groups = 'drop')

# 5. Mean Vitamin D values per PLZ mapping before vs during
mean_vitd_by_plz <- df %>%
  group_by(plz_mapping, period) %>%
  summarise(mean_value = mean(value_result, na.rm = TRUE), .groups = 'drop')

# 6. Mean Vitamin D values per PLZ mapping and age group before vs during
mean_vitd_by_plz_age <- df %>%
  group_by(plz_mapping, age_group, period) %>%
  summarise(mean_value = mean(value_result, na.rm = TRUE), .groups = 'drop')

# 7. Mean Vitamin D values per PLZ mapping and gender before vs during
mean_vitd_by_plz_gender <- df %>%
  group_by(plz_mapping, gender, period) %>%
  summarise(mean_value = mean(value_result, na.rm = TRUE), .groups = 'drop')

# 8. Deficiency rates per PLZ mapping before vs during
def_rate_by_plz <- df %>%
  group_by(plz_mapping, period) %>%
  summarise(deficient = sum(deficient == 1, na.rm = TRUE),
            total = n(),
            deficiency_rate = deficient / total * 100,
            .groups = 'drop')

# 9. Deficiency rates per PLZ mapping and age group before vs during
def_rate_by_plz_age <- df %>%
  group_by(plz_mapping, age_group, period) %>%
  summarise(deficient = sum(deficient == 1, na.rm = TRUE),
            total = n(),
            deficiency_rate = deficient / total * 100,
            .groups = 'drop')

# 10. Deficiency rates per PLZ mapping and gender before vs during
def_rate_by_plz_gender <- df %>%
  group_by(plz_mapping, gender, period) %>%
  summarise(deficient = sum(deficient == 1, na.rm = TRUE),
            total = n(),
            deficiency_rate = deficient / total * 100,
            .groups = 'drop')


print(patients_per_postal_code)
print(patients_per_plz_mapping)
print(patients_by_age_plz)
print(patients_by_gender_plz)
print(mean_vitd_by_plz)
print(mean_vitd_by_plz_age)
print(mean_vitd_by_plz_gender)
print(def_rate_by_plz)
print(def_rate_by_plz_age)
print(def_rate_by_plz_gender)

```

```{r}
# Exact p values

# Formatter
format_p_latex <- function(p, digits = 2) {
  if (length(p) == 0 || is.na(p)) return("$p = \\text{NA}$")
  if (p == 0) return("$p < 1 \\times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)  # e.g. "2.34e-47"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  base <- parts[1]; exp <- as.integer(parts[2])
  paste0("$p = ", base, " \\times 10^{", exp, "}$")
}

# Run Welch t-test per region (plz_mapping = "City"/"Rural area"/"Suburb")
region_pvals <- df %>%
  filter(!is.na(plz_mapping), period %in% c("B","D")) %>%
  group_split(plz_mapping) %>%
  map_dfr(function(subdf) {
    reg <- unique(subdf$plz_mapping)
    if (length(reg) != 1) return(NULL)
    t_res <- tryCatch(
      t.test(value_result ~ period, data = subdf, var.equal = FALSE),
      error = function(e) NULL
    )
    if (is.null(t_res)) {
      tibble(region = reg, p_value = NA_real_, p_formatted_latex = "$p = \\text{NA}$")
    } else {
      tibble(region = reg,
             p_value = t_res$p.value,
             p_formatted_latex = format_p_latex(t_res$p.value))
    }
  })

print(region_pvals)

cat("\nRegion-specific p-values:\n",
    paste0(region_pvals$region, ": ", region_pvals$p_formatted_latex, collapse = "\n"),
    "\n")

# Run χ² test for deficiency per region
region_def_pvals <- df %>%
  filter(!is.na(plz_mapping), period %in% c("B","D")) %>%
  group_split(plz_mapping) %>%
  map_dfr(function(subdf) {
    reg <- unique(subdf$plz_mapping)
    if (length(reg) != 1) return(NULL)
    # contingency table: period x deficiency
    tab <- table(subdf$period, subdf$deficient)
    if (nrow(tab) < 2 || ncol(tab) < 2) {
      tibble(region = reg, p_value = NA_real_, p_formatted_latex = "$p = \\text{NA}$")
    } else {
      chi <- suppressWarnings(chisq.test(tab, correct = FALSE))
      tibble(region = reg,
             p_value = chi$p.value,
             p_formatted_latex = format_p_latex(chi$p.value))
    }
  })

print(region_def_pvals)

cat("\nRegion-specific deficiency p-values:\n",
    paste0(region_def_pvals$region, ": ", region_def_pvals$p_formatted_latex, collapse = "\n"),
    "\n")

```


#### 9.1.1 PSM
```{r}
# Postal Code
# performing exact matching parameters on the entire sample
test_match_obj <- matchit(period ~ age_group + gender + test_month + postal_code, data = df, method = "nearest", distance = "glm", exact = c("gender", "age_group", "test_month", "postal_code")) 

matched_data <- match.data(test_match_obj)

# Calculate the sample sizes
original_size <- nrow(df)  # Total original sample size
matched_size <- nrow(matched_data)  # Total matched sample size

# Calculate the number and percentage of observations lost
num_lost <- original_size - matched_size
percent_lost <- (num_lost / original_size) * 100

# Print results
cat("Original Sample Size: ", original_size, "\n")
cat("Matched Sample Size: ", matched_size, "\n")
cat("Number of Observations Lost: ", num_lost, "\n")
cat("Percentage Lost: ", percent_lost, "%\n")

# Summary
summary(test_match_obj)
```

```{r}
# Calculate summaries for Age with Chi-squared Test for Independence
age_summary <- matched_data %>%
  group_by(period, age_group) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
age_table <- xtabs(count ~ period + age_group, data = age_summary)
age_chi_sq_test <- tryCatch(chisq.test(age_table), error = function(e) e)
age_p_value <- ifelse(class(age_chi_sq_test) == "try-error", NA, age_chi_sq_test$p.value)

# Calculate summaries for Gender with Chi-squared Test
gender_summary <- matched_data %>%
  group_by(period, gender) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
gender_table <- xtabs(count ~ period + gender, data = gender_summary)
gender_chi_sq_test <- tryCatch(chisq.test(gender_table), error = function(e) e)
gender_p_value <- ifelse(class(gender_chi_sq_test) == "try-error", NA, gender_chi_sq_test$p.value)

# Calculate summaries for Test month with Chi-squared Test
test_month_summary <- matched_data %>%
  group_by(period, test_month) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
test_month_table <- xtabs(count ~ period + test_month, data = test_month_summary)
test_month_chi_sq_test <- tryCatch(chisq.test(test_month_table), error = function(e) e)
test_month_p_value <- ifelse(class(test_month_chi_sq_test) == "try-error", NA, test_month_chi_sq_test$p.value)

# Calculate summaries for Season with Chi-squared Test
season_summary <- matched_data %>%
  group_by(period, season) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

# Conduct Chi-squared Test and append the result
season_table <- xtabs(count ~ period + season, data = season_summary)
season_chi_sq_test <- tryCatch(chisq.test(season_table), error = function(e) e)
season_p_value <- ifelse(class(season_chi_sq_test) == "try-error", NA, season_chi_sq_test$p.value)

# Calculate mean and standard deviation for Vitamin D levels and perform t-test
vitamin_d_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    mean_vitamin_d = mean(value_result, na.rm = TRUE),
    sd_vitamin_d = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )
vitamin_d_t_test <- t.test(value_result ~ period, data = matched_data)
vitamin_d_p_value <- vitamin_d_t_test$p.value

# Calculate the percentage of deficiencies for each period and perform Chi-squared Test
deficiency_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    total = n(),
    deficient_count = sum(deficient == 1, na.rm = TRUE),
    non_deficient_count = sum(deficient == 0, na.rm = TRUE),
    deficiency_rate = deficient_count / total * 100,
    .groups = 'drop'
  )

# Create a contingency table for the Chi-squared test
deficiency_table <- xtabs(~ period + deficient, data = matched_data)

# Conduct Chi-squared Test and append the result
deficiency_chi_sq_test <- tryCatch(chisq.test(deficiency_table), error = function(e) e)
deficiency_p_value <- ifelse(class(deficiency_chi_sq_test) == "try-error", NA, deficiency_chi_sq_test$p.value)

# Print the summaries and p-values
print(age_summary)
print(paste("Age Group P-value:", age_p_value))
print(gender_summary)
print(paste("Gender P-value:", gender_p_value))
print(test_month_summary)
print(paste("Test Month P-value:", test_month_p_value))
print(season_summary)
print(paste("Season P-value:", season_p_value))
print(vitamin_d_summary)
print(paste("Vitamin D Levels P-value:", vitamin_d_p_value))
print(deficiency_summary)
print(paste("Deficiency Rate P-value:", deficiency_p_value))
```

```{r}
# Effect Size

# Calculate means and SDs for matched data
mean_matched_b <- mean(matched_data$value_result[matched_data$period == "B"])
mean_matched_d <- mean(matched_data$value_result[matched_data$period == "D"])
sd_matched_b <- sd(matched_data$value_result[matched_data$period == "B"])
sd_matched_d <- sd(matched_data$value_result[matched_data$period == "D"])

# Pooled SD for matched data
n1_matched <- nrow(matched_data[matched_data$period == "B", ])
n2_matched <- nrow(matched_data[matched_data$period == "D", ])
sd_matched_pooled <- sqrt(((n1_matched - 1) * sd_matched_b^2 + 
                           (n2_matched - 1) * sd_matched_d^2) / 
                          (n1_matched + n2_matched - 2))

# SMD (flipped sign for "During - Before")
smd <- (mean_matched_d - mean_matched_b) / sd_matched_pooled  # Flips the direction
print(paste("Standardized Mean Difference (SMD):", round(smd, 4)))

# Standard error for SMD (same formula as Cohen's d)
se_smd <- sqrt((n1 + n2) / (n1 * n2) + (smd^2) / (2 * (n1 + n2)))

# Confidence intervals
ci_99_lower_smd <- smd - z_99 * se_smd
ci_99_upper_smd <- smd + z_99 * se_smd

ci_95_lower_smd <- smd - z_95 * se_smd  
ci_95_upper_smd <- smd + z_95 * se_smd  

# Note: Confidence intervals for the Standardized Mean Difference (SMD) are calculated using the formula:CI = SMD ± Z * SE_SMD where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_SMD is the standard error of the SMD estimate.

# Print results
cat("99% CI for SMD:", ci_99_lower_smd, "to", ci_99_upper_smd, "\n")
cat("95% CI for SMD:", ci_95_lower_smd, "to", ci_95_upper_smd, "\n")  

# Ensure sample sizes are numeric to avoid integer overflow
n1_matched_def <- as.numeric(nrow(matched_data[matched_data$period == "B", ]))
n2_matched_def <- as.numeric(nrow(matched_data[matched_data$period == "D", ]))

# Compute means and standard deviations for matched deficiency rates
mean_matched_b_def <- mean(matched_data$deficient[matched_data$period == "B"])
mean_matched_d_def <- mean(matched_data$deficient[matched_data$period == "D"])
sd_matched_b_def <- sd(matched_data$deficient[matched_data$period == "B"])
sd_matched_d_def <- sd(matched_data$deficient[matched_data$period == "D"])

# Pooled SD for matched deficiency data
sd_matched_pooled_def <- sqrt(((n1_matched_def - 1) * sd_matched_b_def^2 + 
                               (n2_matched_def - 1) * sd_matched_d_def^2) / 
                              (n1_matched_def + n2_matched_def - 2))

# SMD for deficiency (flipped sign for "During - Before")
smd_def <- (mean_matched_d_def - mean_matched_b_def) / sd_matched_pooled_def
print(paste("SMD for deficiency:", round(smd_def, 4)))

# Standard error for SMD
se_smd_def <- sqrt((n1_matched_def + n2_matched_def) / (n1_matched_def * n2_matched_def) + (smd_def^2) / (2 * (n1_matched_def + n2_matched_def)))

# Confidence intervals for SMD
ci_99_lower_smd_def <- smd_def - z_99 * se_smd_def
ci_99_upper_smd_def <- smd_def + z_99 * se_smd_def
ci_95_lower_smd_def <- smd_def - z_95 * se_smd_def
ci_95_upper_smd_def <- smd_def + z_95 * se_smd_def

# Print results
cat("99% CI for SMD:", ci_99_lower_smd_def, "to", ci_99_upper_smd_def, "\n")
cat("95% CI for SMD:", ci_95_lower_smd_def, "to", ci_95_upper_smd_def, "\n")

# Note: Confidence intervals for the Standardized Mean Difference (SMD) are calculated using the formula:CI = SMD ± Z * SE_SMD where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_SMD is the standard error of the SMD estimate.
```

#### 9.1.2 ML

```{r}
# Create treatment variable
df$treatment <- ifelse(df$period == "D", 1, 0)

# One-hot encode categorical variables using model.matrix
X <- model.matrix(~ age_group + gender + season + postal_code, data = df)[, -1]

# Calculate pre-pandemic average
average_value_result <- df %>%
  filter(period == "B") %>%
  summarise(avg_value = mean(value_result, na.rm = TRUE)) %>%
  pull(avg_value)

print(paste("Pre-pandemic average combined value:", round(average_value_result, 4)))

# Run causal forest
causal_forest_model <- causal_forest(
  X = X, 
  Y = df$value_result, 
  W = df$treatment, 
  num.trees = 2000
)

# Estimate ATE
ate_result <- average_treatment_effect(causal_forest_model, target.sample = "all")
ate_estimate <- ate_result[1]
ate_std_err <- ate_result[2]

# Print ATE
print(paste("ATE Estimate:", round(ate_estimate, 4)))

# Calculate % change
percentage_change <- (ate_estimate / average_value_result) * 100
print(paste("Percentage change due to pandemic:", round(percentage_change, 2), "%"))

# Confidence intervals
z_99 <- 2.576
z_95 <- 1.960

ci_99_lower_ate <- ate_estimate - z_99 * ate_std_err
ci_99_upper_ate <- ate_estimate + z_99 * ate_std_err
ci_95_lower_ate <- ate_estimate - z_95 * ate_std_err
ci_95_upper_ate <- ate_estimate + z_95 * ate_std_err

cat("99% CI for ATE:", ci_99_lower_ate, "to", ci_99_upper_ate, "\n")
cat("95% CI for ATE:", ci_95_lower_ate, "to", ci_95_upper_ate, "\n")


# Note: Confidence intervals for the ATE estimate are calculated using the formula: CI = ATE ± Z * SE_ATE where Z is the Z-score for the desired confidence level (e.g., 1.960 for 95%, 2.576 for 99%), and SE_ATE is the standard error of the ATE estimate.
```

#### 9.1.3 FIGURE S4: Plots
```{r}
# Set Z-scores for confidence intervals
z_95 <- qnorm(0.975)  # 1.96
z_99 <- qnorm(0.995)  # 2.576

# Create binary treatment variable
df$treatment <- ifelse(df$period == "D", 1, 0)

# Prepare a data frame to store results
regional_results <- data.frame()

# Loop over each regional group
for (region in unique(df$plz_mapping)) {
  
  df_sub <- df %>% filter(plz_mapping == region)

  # Remove rows with NA in key variables
  df_sub_clean <- df_sub %>%
    filter(
      !is.na(value_result),
      !is.na(treatment),
      !is.na(age_group),
      !is.na(gender),
      !is.na(season),
      !is.na(postal_code)
    )

  # Skip if not enough data
  if (nrow(df_sub_clean) < 100) next

  # Create design matrix with one-hot encoding for covariates (excluding intercept)
  X <- model.matrix(~ age_group + gender + season + postal_code, data = df_sub_clean)[, -1]

  # Run causal forest
  cf <- causal_forest(
    X = X,
    Y = df_sub_clean$value_result,
    W = df_sub_clean$treatment,
    num.trees = 2000
  )

  # Get ATE and SE
  ate <- average_treatment_effect(cf, target.sample = "all")
  ate_estimate <- ate[1]
  ate_se <- ate[2]

  # Calculate confidence intervals
  ci_95_lower <- ate_estimate - z_95 * ate_se
  ci_95_upper <- ate_estimate + z_95 * ate_se
  ci_99_lower <- ate_estimate - z_99 * ate_se
  ci_99_upper <- ate_estimate + z_99 * ate_se

  # Store results
  regional_results <- rbind(regional_results, data.frame(
    Category = paste("Region:", region),
    ATE_Estimate = ate_estimate,
    CI_Lower_95 = ci_95_lower,
    CI_Upper_95 = ci_95_upper,
    CI_Lower_99 = ci_99_lower,
    CI_Upper_99 = ci_99_upper
  ))
}

# Reorder factor levels by ATE estimate for nice plotting
regional_results$Category <- factor(
  regional_results$Category,
  levels = regional_results$Category[order(regional_results$ATE_Estimate)]
)

# Create forest plot
forest_plot <- ggplot(regional_results, aes(x = Category, y = ATE_Estimate)) +
  geom_point(size = 5, shape = 21, fill = "black", color = "black", stroke = 0.5) +
  geom_linerange(aes(ymin = CI_Lower_99, ymax = CI_Upper_99), size = 1.0, color = "black") +
  geom_linerange(aes(ymin = CI_Lower_95, ymax = CI_Upper_95), size = 2.0, color = "firebrick") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  coord_flip() +
  labs(
    x = "Region",
    y = "Estimated pandemic effect on vitamin D (µg/l)"
  ) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold")
  )


```


### 9.2 Longitudinal analyses
#### 9.2.1 Last measurement per person - NOT INCLUDED IN THE PAPER
```{r}
# First: identifying all measurements of patients who have follow-ups
df_multiple_measurements <- df_all_measurements %>%
  group_by(patient) %>%
  filter(n() > 1) %>%
  ungroup()

# Check # Patients with follow-ups
n_distinct(df_multiple_measurements$patient)
```
```{r}
# Limit data frame to only contain the last measurement per patient
df_last_measurement <- df_multiple_measurements %>%
  arrange(patient, cohort_year, test_month) %>%
  group_by(patient) %>%
  slice_tail(n = 1) %>%
  ungroup()
```

```{r}
# Descriptive
sample_size_period <- df_last_measurement %>% count(period)
print(sample_size_period)

sample_size_contingency <- with(df_last_measurement, table(period))
sample_size_p_value <- chisq.test(sample_size_contingency)$p.value
print(sample_size_p_value)

age_group_counts <- df_last_measurement %>% group_by(period, age_group) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print(age_group_counts)

age_group_p_value <- chisq.test(with(df_last_measurement, table(period, age_group)))$p.value
print(age_group_p_value)

gender_group_counts <- df_last_measurement %>% group_by(period, gender) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print(gender_group_counts)

gender_group_p_value <- chisq.test(with(df_last_measurement, table(period, gender)))$p.value
print(gender_group_p_value)

season_counts <- df_last_measurement %>% group_by(period, season) %>%
  summarise(n = n(), .groups = 'drop') %>% group_by(period) %>%
  mutate(percentage = n / sum(n) * 100) %>% ungroup()
print(season_counts)

season_counts_p_value <- chisq.test(with(df_last_measurement, table(period, season)))$p.value
print(season_counts_p_value)

mean_vitamin_d_levels <- df_last_measurement %>% group_by(period) %>%
  summarise(mean_value = mean(value_result, na.rm = TRUE),
            sd_value = sd(value_result, na.rm = TRUE),
            .groups = 'drop')
print(mean_vitamin_d_levels)

vitamin_d_t_test <- t.test(value_result ~ period, data = df_last_measurement)
print(vitamin_d_t_test)

vitamin_d_deficiency_rate <- df_last_measurement %>% group_by(period) %>%
  summarise(deficiency_count = sum(deficient), 
            total_count = n(), 
            deficiency_rate = sum(deficient) / n() * 100, 
            .groups = 'drop')
print(vitamin_d_deficiency_rate)

deficiency_contingency <- with(df_last_measurement, table(period, deficient))
deficiency_p_value <- chisq.test(deficiency_contingency)$p.value
print(deficiency_p_value)

```


```{r}
# PSM
# Ensure variables are factors
df_last_measurement$gender <- factor(df_last_measurement$gender)
df_last_measurement$age_group <- factor(df_last_measurement$age_group)
df_last_measurement$period <- factor(df_last_measurement$period)
df_last_measurement$test_month <- factor(df_last_measurement$test_month)
df_last_measurement$season <- factor(df_last_measurement$season)

# Perform exact matching (no postal_code)
test_match_obj <- matchit(period ~ age_group + gender + test_month, 
                          data = df_last_measurement, 
                          method = "nearest", distance = "glm", 
                          exact = c("gender", "age_group", "test_month")) 

matched_data <- match.data(test_match_obj)

# Sample size statistics
original_size <- nrow(df_last_measurement)
matched_size <- nrow(matched_data)
num_lost <- original_size - matched_size
percent_lost <- (num_lost / original_size) * 100

cat("Original Sample Size: ", original_size, "\n")
cat("Matched Sample Size: ", matched_size, "\n")
cat("Number of Observations Lost: ", num_lost, "\n")
cat("Percentage Lost: ", percent_lost, "%\n")

# Matching summary
summary(test_match_obj)
```
```{r}
# Age group summary and chi-squared test
age_summary <- matched_data %>%
  group_by(period, age_group) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

age_table <- xtabs(count ~ period + age_group, data = age_summary)
age_chi_sq_test <- tryCatch(chisq.test(age_table), error = function(e) e)
age_p_value <- ifelse(class(age_chi_sq_test) == "try-error", NA, age_chi_sq_test$p.value)

# Gender summary and chi-squared test
gender_summary <- matched_data %>%
  group_by(period, gender) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

gender_table <- xtabs(count ~ period + gender, data = gender_summary)
gender_chi_sq_test <- tryCatch(chisq.test(gender_table), error = function(e) e)
gender_p_value <- ifelse(class(gender_chi_sq_test) == "try-error", NA, gender_chi_sq_test$p.value)

# Test month summary and chi-squared test
test_month_summary <- matched_data %>%
  group_by(period, test_month) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

test_month_table <- xtabs(count ~ period + test_month, data = test_month_summary)
test_month_chi_sq_test <- tryCatch(chisq.test(test_month_table), error = function(e) e)
test_month_p_value <- ifelse(class(test_month_chi_sq_test) == "try-error", NA, test_month_chi_sq_test$p.value)

# Season summary and chi-squared test
season_summary <- matched_data %>%
  group_by(period, season) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(period) %>%
  mutate(percentage = count / sum(count) * 100)

season_table <- xtabs(count ~ period + season, data = season_summary)
season_chi_sq_test <- tryCatch(chisq.test(season_table), error = function(e) e)
season_p_value <- ifelse(class(season_chi_sq_test) == "try-error", NA, season_chi_sq_test$p.value)

# Vitamin D summary and t-test
vitamin_d_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    mean_vitamin_d = mean(value_result, na.rm = TRUE),
    sd_vitamin_d = sd(value_result, na.rm = TRUE),
    .groups = 'drop'
  )

vitamin_d_t_test <- t.test(value_result ~ period, data = matched_data)
vitamin_d_p_value <- vitamin_d_t_test$p.value

# Deficiency summary and chi-squared test
deficiency_summary <- matched_data %>%
  group_by(period) %>%
  summarise(
    total = n(),
    deficient_count = sum(deficient == 1, na.rm = TRUE),
    non_deficient_count = sum(deficient == 0, na.rm = TRUE),
    deficiency_rate = deficient_count / total * 100,
    .groups = 'drop'
  )

deficiency_table <- xtabs(~ period + deficient, data = matched_data)
deficiency_chi_sq_test <- tryCatch(chisq.test(deficiency_table), error = function(e) e)
deficiency_p_value <- ifelse(class(deficiency_chi_sq_test) == "try-error", NA, deficiency_chi_sq_test$p.value)

# Print summaries and p-values
print(age_summary); print(paste("Age Group P-value:", age_p_value))
print(gender_summary); print(paste("Gender P-value:", gender_p_value))
print(test_month_summary); print(paste("Test Month P-value:", test_month_p_value))
print(season_summary); print(paste("Season P-value:", season_p_value))
print(vitamin_d_summary); print(paste("Vitamin D Levels P-value:", vitamin_d_p_value))
print(deficiency_summary); print(paste("Deficiency Rate P-value:", deficiency_p_value))
```


```{r}
# ML
df_last_measurement$age_group_numeric <- as.integer(factor(df_last_measurement$age_group))
df_last_measurement$gender_numeric <- as.integer(factor(df_last_measurement$gender))
df_last_measurement$season_numeric <- as.integer(factor(df_last_measurement$season))
df_last_measurement$treatment <- ifelse(df_last_measurement$period == "D", 1, 0)

causal_forest_model <- causal_forest(
  X = df_last_measurement[c("age_group_numeric", "gender_numeric", "season_numeric")], 
  Y = df_last_measurement$value_result, 
  W = df_last_measurement$treatment, 
  num.trees = 2000
)

ate_result <- average_treatment_effect(causal_forest_model, target.sample = "all")

ate_estimate <- ate_result[1]
ate_std_err <- ate_result[2]

print(paste("ATE Estimate:", round(ate_estimate, 4)))

average_value_result <- mean(df_last_measurement$value_result, na.rm = TRUE)
percentage_change <- (ate_estimate / average_value_result) * 100
print(paste("Percentage change due to pandemic:", round(percentage_change, 2), "%"))

ci_99_lower_ate <- ate_estimate - z_99 * ate_std_err
ci_99_upper_ate <- ate_estimate + z_99 * ate_std_err
ci_95_lower_ate <- ate_estimate - z_95 * ate_std_err
ci_95_upper_ate <- ate_estimate + z_95 * ate_std_err

cat("99% Confidence Interval for ATE:", ci_99_lower_ate, "to", ci_99_upper_ate, "\n")
cat("95% Confidence Interval for ATE:", ci_95_lower_ate, "to", ci_95_upper_ate, "\n")
```

#### 9.2.2 Within-subject analysis

```{r}
# Exploration of cohorts 
# Count number of tests per patient per period
test_counts <- df_all_measurements %>%
  filter(period %in% c("B", "D")) %>%
  group_by(patient, period) %>%
  summarise(n_tests = n(), .groups = "drop")

# Create frequency table: how many patients had X tests in each period
test_distribution <- test_counts %>%
  group_by(period, n_tests) %>%
  summarise(n_patients = n(), .groups = "drop") %>%
  arrange(period, n_tests)

# Print the distribution table
print(test_distribution)
```


```{r}
# Keep only patients with measurements in both periods
patients_before <- df_all_measurements %>% filter(period == "B") %>% distinct(patient)
patients_during <- df_all_measurements %>% filter(period == "D") %>% distinct(patient)
 
patients_longitudinal <- inner_join(patients_before, patients_during, by = "patient")
 
df_longitudinal <- df_all_measurements %>%
  semi_join(patients_longitudinal, by = "patient")
 
df_longitudinal$patient <- as.factor(df_longitudinal$patient)
df_longitudinal$period <- factor(df_longitudinal$period, levels = c("B", "D"))
 
model <- lme(
  value_result ~ period + age + gender + season,
  random = ~ period | patient,
  data = df_longitudinal,
  method = "REML"  
)
 
summary(model)
```
 
 
 
### 9.3 Regression analysis add-on

#### 9.3.1 Multivariable logistic regression vitamin D deficiency

```{r}
# Convert variables to factors
df$period <- factor(df$period)
df$age_group <- factor(df$age_group)
df$gender <- factor(df$gender)
df$season <- factor(df$season)
df$deficient <- as.integer(df$deficient)

# Logistic regression model
logistic_model <- glm(deficient ~ period + age_group + gender + season,
                      data = df, family = binomial)

# Summary and odds ratios with 99% confidence intervals
summary(logistic_model)
odds_ratios <- exp(coef(logistic_model))
conf_int_99_logit <- exp(confint(logistic_model, level = 0.99))
print(data.frame(OddsRatio = odds_ratios,
                 CI_lower = conf_int_99_logit[, 1],
                 CI_upper = conf_int_99_logit[, 2]))
```

#### 9.3.2 Multivariable linear regression mean vitamin D levels
```{r}
# Linear regression model
linear_model <- lm(value_result ~ period + age_group + gender + season, data = df)

# Summary and adjusted differences
summary(linear_model)

# 99% confidence intervals for coefficients
conf_int_99_lm <- confint(linear_model, level = 0.99)
print(conf_int_99_lm)

```

### 9.4 Linear Time Trend Analysis

```{r}
# --- p-value formatter -> "p = 5.6 × 10⁻¹⁸" (Unicode superscripts) ----------
format_p <- function(p, digits = 2) {
  sup <- c("0"="⁰","1"="¹","2"="²","3"="³","4"="⁴","5"="⁵",
           "6"="⁶","7"="⁷","8"="⁸","9"="⁹","-"="⁻","+"="⁺")
  to_sup <- function(n) paste0(sup[strsplit(as.character(n), "")[[1]]], collapse = "")
  vapply(p, function(x){
    if (is.na(x)) return(NA_character_)
    s <- formatC(x, format = "e", digits = digits)   # e.g., "5.60e-18"
    parts <- strsplit(s, "e")[[1]]
    paste0("p = ", parts[1], " × 10", to_sup(as.integer(parts[2])))
  }, character(1))
}

# ============================================================================
# Create dummy variables for predictors (excluding intercept)
X_trend <- model.matrix(~ age_group + gender + season, data = df)[, -1]
colnames(X_trend) <- make.names(colnames(X_trend))  # Clean names

# Combine into modeling data frame
df_trend <- cbind(value_result = df$value_result,
                  deficient    = df$deficient,
                  time_numeric = df$time_numeric,
                  X_trend)

# Build formulas dynamically
predictors   <- colnames(df_trend)[!(colnames(df_trend) %in% c("value_result", "deficient"))]
formula_lm   <- as.formula(paste("value_result ~", paste(predictors, collapse = " + ")))
formula_logit<- as.formula(paste("deficient ~",     paste(predictors, collapse = " + ")))

# Fit models
lm_time    <- lm(formula_lm, data = as.data.frame(df_trend))
logit_time <- glm(formula_logit, data = as.data.frame(df_trend), family = binomial)

# Tidy model outputs with 99% CI + formatted exact p-values
lm_99ci <- broom::tidy(lm_time, conf.int = TRUE, conf.level = 0.99) %>%
  dplyr::mutate(p_value_exact = format_p(p.value)) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p_value_exact)

logit_99ci <- broom::tidy(logit_time, conf.int = TRUE, conf.level = 0.99) %>%
  dplyr::mutate(p_value_exact = format_p(p.value)) %>%
  dplyr::select(term, estimate, conf.low, conf.high, p_value_exact)

# Print results
cat("\n=== Linear Model: 99% CI + exact p-values ===\n")
print(lm_99ci)

cat("\n=== Logistic Model: 99% CI + exact p-values ===\n")
print(logit_99ci)

```

```{r}
# Exact p values time variable

# LaTeX-ready p-value formatter (single backslash)
format_p_latex <- function(p, digits = 2) {
  if (is.na(p)) return("$p = \\text{NA}$")
  if (p == 0)   return("$p < 1 \\times 10^{-308}$")
  s <- formatC(p, format = "e", digits = digits)  # e.g. "2.34e-47"
  parts <- strsplit(s, "e", fixed = TRUE)[[1]]
  paste0("$p = ", parts[1], " \\times 10^{", as.integer(parts[2]), "}$")
}

# 1) Build continuous time in months from test_date
anchor_date <- as.Date("2018-03-01")
df <- df %>%
  mutate(
    test_date   = as.Date(test_date),
    time_months = as.numeric(interval(anchor_date, test_date) / months(1))
  )

# 2) Fit models (time + covariates)
lm_time    <- lm(value_result ~ time_months + age_group + gender + season, data = df)
logit_time <- glm(deficient    ~ time_months + age_group + gender + season, data = df, family = binomial)

# 3) Tidy, vectorized formatting of p-values
lm_99ci <- tidy(lm_time, conf.int = TRUE, conf.level = 0.99) %>%
  mutate(p_value_latex = map_chr(p.value, format_p_latex))

logit_99ci <- tidy(logit_time, conf.int = TRUE, conf.level = 0.99) %>%
  mutate(p_value_latex = map_chr(p.value, format_p_latex))

# Extract LaTeX-ready p-values for Time (months)
p_time_linear_latex <- lm_99ci  %>% filter(term == "time_months")  %>% pull(p_value_latex) %>% .[1]
p_time_logit_latex  <- logit_99ci %>% filter(term == "time_months") %>% pull(p_value_latex) %>% .[1]

cat(
  "\nExact p-values for Time (months):\n",
  "Linear model:  ", p_time_linear_latex,  "\n",
  "Logistic model:", p_time_logit_latex,   "\n",
  sep = ""
)

```


